PIM kinase control of CD8 T cell protein synthesis and cell trafficking

Version 1

Abstract

Abstract

Integration of a large network of kinase signalling pathways co-ordinates changes in the transcription, translation and metabolic events required for T cell activation and differentiation. The present study explores the role of the Serine/Threonine kinases PIM1 and PIM2 in controlling murine CD8 T lymphocyte antigen receptor-mediated activation and differentiation in response to the cytokines Interleukin 2 (IL-2) or IL-15. We show that PIM kinases are dispensable for the differentiation programs controlled by the antigen-receptor and IL-15. There is however a selective role for the PIM kinases in the context of IL-2 regulation of CD8 T cell fate. One key insight was that the PIM kinases controlled the migratory capabilities of effector CD8 T cells, with Pim1/Pim2-deficient CD8 T cells unable to fully switch off the naïve T cell chemokine and adhesion receptor program during effector differentiation. PIM kinases were also needed for IL-2 to sustain high expression of the glucose transporters SLC2A1 and SLC2A3 and to maintain activity of the nutrient sensing kinase mTORc1. Strikingly, PIM kinases did not have a dominant impact on IL-2-driven transcriptional programs but rather selectively modulated protein synthesis to shape cytotoxic T cell proteomes. This study reveals a selective role of PIM kinases in IL-2 control of CD8 T cells and highlights how regulated changes in protein synthesis can impact T cell phenotypes.

Introduction

CD8 T lymphocytes are key effector cells for adaptive immune responses to viruses, intracellular pathogens and tumours. Their clonal expansion and differentiation to cytotoxic effector or memory CD8 T cells is controlled by signalling pathways initiated by the T cell antigen receptor (TCR), co-stimulatory molecules and cytokines. Receptor engagement by immune stimuli triggers the activation of cytosolic tyrosine kinases that then direct the activity of a diverse network of Serine/Threonine kinases, disseminating signals from plasma membrane to cell interior. The Src family and ZAP70 Tyrosine kinases initiate this process downstream of the TCR, with the Janus kinases (JAK) as the main signal transducer for important cytokine regulators of T cell differentiation, IL-2 and IL-15 (Cantrell, 2015; Saravia et al., 2020). The subsequent network of Serine/Threonine kinases that is engaged beyond these initiating events coordinates phosphorylation of a diverse array of protein targets.

These protein targets include chromatin regulators, transcription factors, translation machinery, cytoskeletal proteins and metabolic enzymes, which then control the changes in transcriptional, translational and metabolic programs that drive effector and memory T cell differentiation (Navarro et al., 2011; Tan et al., 2017; Locard-Paulet et al., 2020).

CD8 T cells express ∼200 Serine/Threonine protein kinases (Navarro et al., 2014; Howden et al., 2019). There is a good understanding of the roles of some Serine/Threonine kinases in CD8 T cells including the role of the diacylglycerol regulated kinases, MAP kinases ERK1 and 2, Phosphatidylinositol (3,4,5) triphosphate-controlled kinases PDK1 and AKT, and the nutrient sensing kinases mammalian Target of Rapamycin complex 1 (mTORc1) and AMP-activated protein kinase (AMPK) (Finlay et al., 2012; Rolf et al., 2013; Blagih et al., 2015; Cantrell, 2015; Hukelmann et al., 2016; Tan et al., 2017; Howden et al., 2019; Saravia et al., 2020; Damasio et al., 2021; Spinelli et al., 2021). However, beyond these core kinases there is much still to learn about how many of the ∼200 Serine/Threonine kinases expressed in CD8 T cells contribute to the control of T cell function. In this context, the PIM kinases are a family of Serine/Threonine kinases that are strongly and rapidly induced in T cells responding to immune stimuli (Aho et al., 2005; Fox et al., 2005; Peperzak et al., 2010; Jackson et al., 2012). The PIM family contains 3 Serine/Threonine kinases, PIM1, PIM2 and PIM3, of which T lymphocytes predominantly express PIM1 and PIM2 (Mikkers et al., 2004; Fox et al., 2005). As PIM kinases are constitutively active, their participation in signalling cascades is dictated by regulated changes in their protein expression levels (Warfel and Kraft, 2015). PIM1 and PIM2 have little to no expression in naïve T cells but are induced by T-cell receptor activation and costimulation (e.g. CD27) (Wingett et al., 1996; Fox et al., 2005; Peperzak et al., 2010). PIM kinases are also direct transcriptional targets of JAK/STAT signalling pathways and as such PIM1 and/or PIM2 are rapidly induced in T cells downstream of a diverse array of cytokines, including multiple common cytokine receptor γ-chain family cytokines, IL-6 and IL-12 (Matikainen et al., 1999; Aho et al., 2005; Fox et al., 2005; Warfel and Kraft, 2015; Buchacher et al., 2023).

There have been some analyses of how deletion of PIM kinases impacts CD8 T cell function. A mild reduction in Pim1/Pim2/Pim3-deficient CD4 and CD8 T cell proliferation has been observed in response to sub-optimal TCR stimulation and IL-2 (Mikkers et al., 2004) and there is evidence that PIM1 and PIM2 control survival pathways in naïve CD4 and CD8 T cells in concert with mTORc1 (Fox et al., 2005). Pim1/Pim2/Pim3 deletion is reported to promote expression of stemness and memory-associated transcription factors including Bcl6 and Tcf7 (Chatterjee et al., 2019) and PIM kinase activity has been linked with maintenance of memory CD8 T cell survival (Knudson et al., 2017). It has also been shown recently that Pim1/Pim2 deficiency prevented mTORc1 activation and the proliferation of non-conventional intra-epithelial gut CD8 T lymphocytes (IEL) in response to the cytokine IL-15 (James et al., 2021).

Elevated PIM kinase expression has been noted in patient T cells for a number of autoimmune conditions, including rheumatoid arthritis, uveitis and coeliac disease, and PIM kinase inhibition has been proposed as a possible intervention to dampen autoimmunity (James et al., 2021; Maney et al., 2021; Li et al., 2022). PIM kinase inhibitors have also entered clinical trials to treat some cancers and there is active interest in combining these with immunotherapies (Clements and Warfel, 2022). Given the role of CD8 T cells in anti-cancer immunity, it is important to understand how these kinase control CD8 T cell function. However, there has not yet been a systematic analysis of the role of PIM kinases in conventional CD8 T cells as they activate and differentiate to form effector and memory cells. Accordingly, the current study, uses high resolution quantitative analysis of Pim1/Pim2-deficient CD8 T cell proteomes and transcriptomes to obtain a comprehensive and unbiased understanding of how PIM kinases control CD8 T cell TCR-driven activation, and IL-2 or IL-15-driven proliferation, and differentiation. The study reveals a selective role for PIM kinases as mediators of IL-2 signalling to control CD8 T cell trafficking and effector functionality. We uncover that the dominant impact of PIM kinase loss was to shape effector CD8 T cell proteomes rather than transcriptomes, reflecting PIM kinase modulation of protein synthesis.

Results

PIM1 and PIM2 are strongly induced by antigen receptor engagement but are dispensable for T cell activation

T cell antigen receptor (TCR)-driven activation of naïve T cells initiates their clonal expansion and differentiation into effector and memory cells. Quantitative mass spectrometry-based proteomics analysis of naïve OT1 CD8 T cell stimulated with their cognate peptide over time (Marchingo et al., 2020) showed that PIM1 protein was strongly induced within 3 hours of TCR engagement, which was sustained at a 24 hour time point (Fig 1A). PIM2 expression was not detected until 9 hours post-stimulation but was also sustained at the 24 hour time point. 24 hour polyclonal activation of CD8 T cells with αCD3/αCD28 agonist antibodies similarly induced expression of PIM1 and PIM2 proteins (Fig 1B, Table S1). These increases in PIM kinases expression at a protein level were underpinned by transcriptional reprograming, with both Pim1 and Pim2 mRNA very lowly expressed in naïve CD8 T cells and strongly upregulated by immune activation (Fig 1C) (Spinelli et al., 2021). Consistent with previous reports of Pim1 and Pim2 being the major PIM kinases in T cells (Mikkers et al., 2004; Fox et al., 2005), Pim3 mRNA was only expressed at very low levels in naïve and activated CD8 T cells (Fig 1C), with PIM3 protein not detected at all by mass spectrometry (Table S1).

Pim1 and Pim2 are strongly TCR-induced, but dispensable for T cell activation

Estimated copies per cell of PIM1 and PIM2 protein from quantitative proteomics analysis of (A) OT1 CD8 T cells stimulated with SIINFEKL peptide for indicated times from published dataset (Marchingo et al., 2020) or (B) naïve ex vivo and 24 hour αCD3/αCD28 (TCR) activated WT CD8 T cells (see Fig 1G-H) for further details. (C) Fragments per kilobase million (FPKM) of Pim1, Pim2 and Pim3 mRNA from published (Spinelli et al., 2021) bulk RNAseq analysis of naïve and 24 hour gp33-41 peptide stimulated P14 CD8 T cells. Lymph node cell suspensions from C57BL/6 (WT) and Pim1KO/Pim2KO (Pim dKO) mice were activated for 24 hours with αCD3/αCD28 (both 0.5 µg/mL) and CD4 and CD8 T cell (D) FSC-A SSC-A profiles, (E) expression of surface activation markers (CD25, CD44, CD71) or CD8 T cell intracellular IFNγ were measured by flow cytometry. (F) Lymph node single cell suspensions from WT and Pim dKO mice were labelled with CellTrace Violet (CTV), activated with αCD3/αCD28 (both 0.5 µg/mL) and CD4 and CD8 T cell CTV proliferation profiles were measured at indicated time points. (G-H) Lymph node cell suspensions from WT and Pim dKO mice were stimulated for 24 hours with αCD3/αCD28 (both 0.5 µg/mL) and activated CD4 and CD8 T cells were sorted for analysis by quantitative proteomics. Data was analysed using proteomic ruler method (Wisniewski et al., 2014) to estimate protein copy number per cell. (G) Total protein content (µg/million cells) (one-way ANOVA), (H) Volcano plots of p-value (-log10) versus fold-change (log2) in protein copy number between Pim dKO and WT. Horizontal dotted line represents multi-test correction cut-off of q=0.05, vertical dotted line shows 1.5-fold change. Lymph node single cell suspensions from WT and Pim dKO mice were labelled with CellTrace Violet (CTV) and (I) cells were cultured in IL-7 (5ng/mL) +/-rapamycin (20 nM) and CD8 T cell numbers measured over time or (J) cells were activated with αCD3/αCD28 (both 0.5 µg/mL) +/-rapamycin (20 nM) and CD8 T cell mean division number was calculated over time (two-way ANOVA). Symbols in bar charts represent biological replicates, symbols in (I) represent the mean. Error bars show mean ± S.D. Flow cytometry dot plots and histograms are representative of (D-E) n = 3, except for IFNγ staining which is n=2 or show pooled data from (I) n=3-4 and (F, J) n=5 biological replicates, with data collected over at least 2 independent experiments. Quantitative proteomics was performed on biological triplicates.

PIM1 and PIM2 have similar substrate selectivity and hence functional redundancy (Mikkers et al., 2004; Bullock et al., 2005; Fox et al., 2005; Peng et al., 2007; Warfel and Kraft, 2015). Accordingly, to explore the role of the PIM kinases in T cells, mice with a whole body deletion of Pim1 and Pim2, Pim1-/- (Pim1 KO) (Laird et al., 1993) and Pim2-/- or Pim2-/Y (Pim2 KO) (Mikkers et al., 2004), were backcrossed for >10 generations onto a C57BL/6 background and inter-crossed to generate Pim1/Pim2 double knockout (Pim dKO) mice.

There were normal proportions of peripheral T cells in spleens of Pim dKO mice, though the total number of splenocytes was lower than in age/sex matched wild-type (WT) mice (Fig S1A-C). However, Pim dKO T cells showed a normal increase in cell size in response to immune activation with αCD3/αCD28 agonist antibodies (Fig 1D) and 24 hour αCD3/αCD28 activated Pim dKO CD4 and CD8 T cells exhibited normal upregulation of established markers of T cell activation including CD25, CD44 and the transferrin receptor CD71 (Fig 1E). 24 hour TCR-activated Pim dKO CD8 T cells also produced normal levels of the key effector cytokine IFNγ (Fig 1E) and there was no discernible difference in TCR-driven proliferation between WT and Pim dKO T cells over 3 days in culture (Fig 1F, J). To systematically explore in an unbiased way if Pim1/Pim2-deficiency impacted TCR-driven activation phenotypes, we performed high-resolution mass spectrometry to measure the proteomes of 24 hour αCD3/αCD28 activated WT and Pim dKO CD4 and CD8 T cells. These experiments quantified expression of >7000 proteins but found no substantial quantitative or qualitative differences in protein content or proteome composition in activated WT versus Pim dKO CD4 and CD8 T cells (Fig 1G-H) (Table S1), indicating a redundancy between PIM kinases and other signalling pathways downstream of TCR activation.

One prominent study has reported that Pim1/Pim2 deficiency sensitises T cells to the mTORc1 inhibitor rapamycin (Fox et al., 2005), which suggests mTORc1 as a candidate for mediating this redundancy. This study showed naïve Pim dKO T cells cultured in IL-7 died in response to rapamycin treatment, whereas WT cells did not, and that rapamycin synergised with Pim1/Pim2-deficiency to substantially reduce TCR-driven activation and proliferation (Fox et al., 2005). However, we found no impact of Pim1/Pim2-deficiency on the ability of naïve CD8 T cells cultured with IL-7 to survive in response to rapamycin treatment (Fig 1I). Furthermore, we found no difference in the effect of rapamycin on mean division number in Pim dKO versus WT CD8 T cells (Fig 1J). Collectively these data show that PIM 1 and PIM2 are rapidly and strongly expressed in response to immune activation, but they are dispensable for the antigen receptor-driven proteome remodeling that initiates T cell clonal expansion and differentiation. We also found no evidence that the Pim1 and Pim2 loss synergised with mTORC1 inhibition.

PIM kinases do not control IL-15-driven CD8 T cell memory differentiation

The proliferative expansion and differentiation programs of CD8 T cells are shaped by cytokines. In this context, PIM1 and PIM2 kinase expression is regulated by the cytokines IL-2 and IL-15 which are key to shaping CD8 T cell effector and memory cell differentiation respectively (Weninger et al., 2001; Cornish et al., 2006; Buck et al., 2016). Both IL-2 and IL-15 control PIM1 and PIM2 expression in antigen primed CD8 T cells (Fig 2A). Accordingly, we assessed the contribution of PIM1 and PIM2 kinases to cytokine-driven expansion and differentiation of antigen activated CD8 T cells. In these experiments, antigen-primed CD8 T cells are expanded in IL-15 to generate memory phenotype CD8 T cells or expanded in IL-2 to generate effector cytotoxic T lymphocytes (CTL) (Weninger et al., 2001; Cornish et al., 2006; Buck et al., 2016) (Fig 2B).

Pim1/Pim2 deficiency reduces IL-15-driven CD8 T cell proliferation but does not prevent memory differentiation.

(A) OT1 lymph node cell suspensions were SIINFEKL peptide activated for 36 hours, washed then cultured with no cytokine, IL-15 (20 ng/mL) or IL-2 (20 ng/mL) for 4 or 24 hours. Western blots of PIM1, PIM2 or pSTAT5 Y694 expression. (B) Schematic of cytokine driven memory and effector CD8 T cell expansion and differentiation cultures. Lymph node or spleen cell suspensions were activated for 2 days with TCR stimulus + cytokine, washed, then split daily into fresh media + cytokine. (C) WT (Ly5.1) and Pim dKO LN suspensions were mixed at a 50:50 ratio for T cells and cultured as outlined in (B) with TCR stimulus αCD3/αCD28 (both 0.5 µg/mL) + cytokine IL-15 (20 ng/mL), and CD8 T cell number was measured daily. (D) WT and Pim dKO T cells were expanded with IL-15 in separate cultures as per (B-C) and % live cells (PI-ve) were assessed on days 4 and 6 (two-way ANOVA). (E-J) WT and Pim dKO CD8 T cells were activated with TCR stimulus αCD3/αCD28 (both 0.5 µg/mL) + cytokine IL-15 (20 ng/mL), expanded with IL-15 as per (B), with an additional CD4 T cell magnetic depletion step on day 3 of culture. CD8 T cells were harvested on day 6 for parallel RNAseq and proteomic analysis. (E) Fold-change in mRNA expression between Pim dKO and WT versus average mRNA expression (TPM). mRNA expression (Transcripts per million, TPM) of (F) secondary lymphoid homing receptors Sell, Ccr7, S1pr1 and (G) key transcription factors involved in CD8 T cell memory differentiation and maintenance Tcf7, Klf2, Foxo1, Foxo3, Id3. (H) WT vs Pim dKO protein copy numbers, differentially expression proteins (FC>1.5, q<0.05) are highlighted in red (I) Protein copy numbers per cell for key mitochondrial proteins DRP1, OPA1 and CPT1A. (J) WT vs Pim dKO protein copy numbers, mitochondrial proteins (as defined in MitoCarta 3.0) are highlight in pink. Symbols in bar charts represent biological replicates, symbols in (C,H,J) represent the mean. Error bars show mean ± S.D. Data are representative of (A) n=3 or show pooled data from (C) n = 4, and (D) n=5 biological replicates with data collected over at least 2 independent experiments. Quantitative proteomics and RNAseq was performed on biological triplicates. ** q:≤0.01, fold-change (FC) shown on bar graphs when q<0.05.

Previous work has shown that PIM1 and PIM2 mediate IL-15 induced proliferation of gut intra-epithelial lymphocytes (IEL) (James et al., 2021). However, their role in IL-15 responses in conventional antigen receptor activated CD8 T cells is not known. We therefore first examined the impact of Pim1/Pim2-deficiency on IL-15-driven conventional CD8 T cell expansion. Antigen experienced WT CD8 T cells show a robust proliferative response to IL-15 such that there is an ∼90-fold expansion of the CD8 T cells between days 2 and 6 of culture (Fig 2C). There was no impact of Pim1/Pim2-deficiency on the ability of IL-15 to sustain CD8 T cell viability (Fig 2D). Immune activated Pim dKO CD8 T cells also proliferated strongly in response to IL-15, but showed a moderate proliferative disadvantage resulting in a ∼4-fold difference in total cell number by day 6 of the culture (Fig 2C). Antigen-primed CD8 T cells expand in IL-15 to generate memory phenotype CD8 T cells (Weninger et al., 2001).

We therefore examined whether Pim1/Pim2-deficiency had any impact on the ability of IL-15 to induce memory differentiation. Accordingly, we performed parallel bulk RNAseq and high-resolution mass spectrometry to analyse the transcriptome and proteome of IL-15 expanded WT and Pim dKO CD8 T cells. The RNAseq analysis quantified ∼14,000 RNA and using a cut off of >1.5 fold-change and q-value <0.05 we saw that the abundance of 381 RNA was modified by Pim1/Pim2-deficiency (Fig 2E) (Table S2A). Of the 381 RNA that were detected as differentially expressed (FC>1.5, q<0.05) only ∼30% were detected with an average expression over 10 transcripts per million (TPM), of these gene ∼36% were non-coding RNA, uncharacterised genes or pseudogenes, leaving only 72 substantially expressed protein coding mRNA that were differentially expressed in Pim dKO versus WT IL-15 expanded CD8 T cells (Table S2B). These data indicate that Pim1/Pim2-deficiency has very little effect on IL-15 induced protein-coding mRNA expression in differentiated T cells. We then specifically examined if Pim dKO T cells had any differences in mRNA that are known to be critical for memory cell differentiation. We found that Pim1/Pim2-deficiency did not decrease expression of mRNA levels for critical memory T cell genes such as the adhesion molecules and chemokine receptors, Sell (CD62L), Ccr7 and S1pr1 that direct the trafficking of CD8 T cells to secondary lymphoid tissue. (Fig 2F). Pim1/Pim2-deficiency also did not reduce expression of mRNA for the key memory cell transcription factors Tcf7, Foxo1, Foxo3 Klf2 and Id3. (Fig 2G). Indeed, if anything, there were small increases in expression of the Klf2, Id3 and Ccr7 mRNA in Pim dKO versus WT IL-15 expanded CD8 T cells (Fig 2F-G).

Critically, when the proteomes of IL15 expanded WT and Pim dKO CD8 T cells were analysed and expression of ∼6900 total proteins quantified, there were no major changes in the proteomes between WT and Pim dKO IL15 expanded CD8 T cells (Fig 2H, Table S3). Of note, the mitochondrial proteome composition of IL15 expanded WT and Pim dKO CD8 T cells were indistinguishable (Fig 2I-J). This is pertinent because one key feature of how IL-15 controls memory T cells is via mitochondrial remodelling to support energy production via oxidative phosphorylation and fatty acid oxidation (Buck et al., 2016) and it has been described that PIM kinases can control mitochondrial phenotypes in cancer cell lines and cardiomyocytes (Din et al., 2013; Chauhan et al., 2020). These data show that PIM1 and PIM2 do not drive IL-15-mediated metabolic or differentiation programs in antigen-primed CD8 T cells.

PIM kinases selectively modulate IL-2 controlled effector CD8 T cell differentiation

IL-2 controls the expression of PIM1 and PIM2 kinases (Fig 2A) and in this context, we noted that the copy numbers per cell of PIM1 protein is considerably higher in actively expanding antigen-activated CD8 cells maintained with IL-2 compared to IL-15 (Fig 3A). We therefore examined the role of PIM1 and PIM2 in the IL-2 signalling pathways that control the differentiation of effector CTL. The data show that Pim dKO antigen activated CD8 T cells responded normally to IL-2 to induce STAT5 Y694 phosphorylation (Fig 3B). They also express normal levels of the IL-2-receptor alpha chain, CD25, (Fig 3C), which is sustained by STAT5, and together these data broadly indicate IL-2 receptor signal transduction is unaltered by Pim1/Pim2-deficiency. Antigen activated WT CD8 T cells have a strong survival and proliferative response to IL-2 producing an ∼270-fold expansion between days 2 and 6 of culture in saturating IL-2 concentrations (Fig 3D). There was no impact of Pim1/Pim2-deficiency on the ability of IL-2 to maintain CD8 T cell viability (Fig 3E) and Pim dKO cells CD8 T cells cultured in IL-2 proliferated robustly. They did however exhibit a very mild proliferative disadvantage resulting in a ∼2-fold difference in total cell number by day 6 of the culture (Fig 3D). We then examined whether Pim1/Pim2-deficiency had any impact on the differentiation of Pim dKO CD8 T cells into effector CTL. Expression of activation markers CD44 and CD71 in IL-2-expanded CTL were similar between WT and Pim dKO CD8 T cells on days 3 and 6 of co-cultures (Fig 3F). However, one consistent difference between IL-2 maintained WT and Pim dKO CD8 T cells is that WT CTL downregulated expression of CD62L (L-selectin, Sell), whereas the Pim dKO CTL sustained higher CD62L expression (Fig 3G). We also noted that IL-2 maintained Pim dKO CTL had reduced forward and side scatter properties i.e were smaller and less granular than corresponding WT CTL (Fig 3F).

Pim dKO IL-2 differentiated effector T cells have reduced cell size and sustained expression of CD62L

(A) Estimated copies per cell of PIM1 and PIM2 protein from published quantitative proteomics analysis (Howden et al., 2019; Brenes et al., 2023) of CD8 T cells expanded in IL-2 or IL-15 as outlined in Fig 2B. (B-D, F-G) WT (Ly5.1) and Pim dKO lymph node or spleen single cell suspensions were mixed at a 50:50 ratio of T cells, activated for 2 days with αCD3/αCD28 (both 0.5 µg/mL) and IL-2 (20 ng/mL), washed then split into fresh medium containing IL-2 (20 ng/mL) daily (as per Fig 2B). Some of the mixed cell suspensions were also cultured in IL-7 (5 ng/mL) to sustain a naïve T cell reference. (B) WT and Pim dKO CTL were treated 1 hour +/-Jak1/3 inhibitor Tofacitinib (100 nM) (negative control) before pSTAT5 Y694 expression was measured on day 3 and 6 of culture, (C) surface CD25 expression was measured on days 3 and 6 or culture, (D) CD8 T cell number vs time was calculated, (F) CD8 T cell FSC-A, SSC-A and surface activation markers (CD44, CD71) were measured on days 3 and 6 of culture (G) expression of adhesion molecule CD62L was measured daily. (E) WT and Pim dKO T cells were activated and expanded with IL-2 as per (Fig 3B-D, F-G) except in separate cultures and % live cells (PI-ve) was assessed on days 4 and 6 (two-way ANOVA). Symbols in bar charts represent biological replicates, symbols in (D) represent the mean. Error bars show mean ± S.D. Data are representative of (B, G) n=4, (C,F) n=6 or show pooled data from (D) n=4, (E) n=6 biological replicates with data collected over at least 2 independent experiments

To explore the apparent size change between Pim dKO and WT IL-2 expanded CTL, we used high resolution mass spectrometry to compare the proteomes of day 6 IL-2 cultured WT or Pim dKO CTL. These analyses quantified expression of ∼6,900 proteins and showed a small decrease in the total protein mass of IL-2 maintained Pim dKO CD8 T cells compared to WT cells (Fig 4A, Table S4). The data showed down-regulation of the expression of 513 proteins (>1.5 fold-change, q<0.05) in Pim dKO CTL along with increased copy numbers of 33 proteins (Fig 4B, Table S4). The transcription factor profile of Pim dKO T cells was consistent with that of an effector CTL with high levels of TBX21 and low levels of TCF1 (Tcf7) (Fig 4C). Pim dKO T cells also exhibited an effector-like mitochondrial proteome profile (Fig S2). However, there were salient differences in proteins important for effector functionality between WT and Pim dKO CD8 T cells. These included lower levels of the glucose transporters, SLC2A1 and SLC2A3 (GLUT1 and GLUT3) (Fig 4D) which are the major glucose transporters in CD8 T cells (Hukelmann et al., 2016) and are key for enabling T cells to fuel high rates of glycolysis and protein O-GlcNAcylation required to support effector functions(Chang et al., 2015; Swamy et al., 2016). Pim dKO CTL also had decreased expression of a number of proteins involved in fatty acid and cholesterol biosynthesis including SCD2, HMGCS1, MVK, IDI1, FDPS, FDFT1 and SQLE (Fig 4E). Another striking observation was that the Pim dKO CD8 cells had reduced levels of a number of key effector proteins including multiple granzymes, perforin, IFNγ and TNFα (Fig 4F). Orthogonal flow cytometric analysis of Granzyme B and IFNγ levels confirmed reduced levels of these effector molecule in IL-2 expanded Pim dKO CTL, though IFNγ protein is only lowly produced in IL2 maintained CTL. (Fig 4G). In parallel experiments we also examined the impact of an acute loss of PIM kinase activity on the proteomes of IL-2 maintained CTL to see if this could recapitulate key observations from Pim dKO T cells. When IL-2 maintained WT CTL were treated for 24 hours with the pan-PIM kinases inhibitors PIM447 or AZD1208 we saw reductions in cell proliferation, cell size, cell granularity, fatty acid biosynthesis proteins SCD1-3, glucose transporters SLC2A1, SLC2A3 and a reduction in Granzyme B expression (Fig S3A-E, Table S5). This corroborated some key observations from the experiments with Pim dKO CTL, where Pim1 and Pim2 had been deleted over the lifespan of the T cell.

Major glucose transporters and effector proteins are reduced in Pim dKO IL-2 expanded CTL

WT and Pim dKO CD8 T cells were activated for 2 days with αCD3/αCD28 (both 0.5 µg/mL) and IL-2 (20 ng/mL), washed then split into fresh medium containing IL-2 (20 ng/mL) daily (as per Fig 2B), with an additional CD4 T cell magnetic depletion step on day 3 of culture. CD8 T cells were harvested on day 6 of culture for high resolution mass spectrometry. (A) Estimated total protein content per cell (student t-test). (B) Volcano plots Pim dKO vs WT protein copy numbers, differentially expressed proteins (FC>1.5, q<0.05) are highlighted in red. Estimated protein copy number per cell of (C) transcription factor TBX21 and TCF1 (D) glucose transporters SLC2A1 and SLC2A3. (E) Volcano plots Pim dKO vs WT protein copy numbers. Proteins with KEGG term = “terpenoid backbone biosynthesis”, “biosynthesis of unsaturated fatty acids” or “steroid biosynthesis” are highlighted with proteins with FC>1.5, q<0.05 shown in red and proteins with FC<1.5 and/or q>0.05 shown in pink. (F) Heatmap of protein copy numbers for granzymes, perforin and effector cytokines. (G) Granzyme B and IFNγ expression was measured by flow cytometry in day 6 IL-2 expanded WT and Pim dKO CTL. Symbols in bar charts show biological replicates. Error bars show mean ± S.D. Data are representative of (G) n = 3-4, with data collected over at least 2 independent experiment Quantitative proteomics was performed on biological triplicates. * indicates q<0.05, fold-change (FC) shown on graph when q<0.05

To examine the molecular regulation underpinning PIM1 and PIM2 kinase control of key metabolic and effector proteins we also performed RNAseq on IL-2 differentiated WT and Pim dKO CTL samples collected in parallel with the proteomics analysis described in Fig 4. In these experiments ∼14,000 mRNA were quantified and using a cut off >1.5 fold-change, q<0.05 we saw that the abundance of 223 mRNA were decreased and 155 increased by Pim1/Pim2 deficiency (Fig 5A, Table S6). These data revealed that the reduced expression of perforin and Granzymes C-G and K protein corresponded to decreases in their mRNA (Fig 5B-C). However, there were striking examples where a clear decrease in protein levels did not correspond with an appreciable decrease in mRNA expression. These included the predominant cytolytic Granzymes B and A (Fig 5C-D), as well as both glucose transporters SLC2A1 and SLC2A3 (Fig 5E). Furthermore, contrasting differential expression of proteins and mRNA data in Pim dKO vs WT CTL revealed that ∼75 % of the proteins whose expression was changed in Pim dKO T cells (FC>1.5, q<0.05), exhibited no strong corresponding changes in their mRNA (FC<1.2) (Fig 5F, Table S7). Indeed, of the ∼500 proteins whose expression decreased at the protein level in Pim dKO CTL only 17 showed a strong decrease in magnitude at the mRNA level as well (Fig 5G, Table S7).

Disconnect between protein and mRNA expression in Pim1/Pim2-deficient effector CTL corresponds with a reduction in protein synthesis

RNAseq analysis was performed in day 6 IL-2 expanded WT and Pim dKO CD8 T cells which were collected in parallel with proteomics analysis described in Fig 4A. (A) Volcano plot of RNAseq data, differentially expressed mRNA (FC>1.5, q<0.05) are highlighted in red. (B) Volcano plot of RNAseq data, Granzymes C-K, perforin, Pdcd4 and Sell are highlighted in red. (C) Heatmap of mRNA expression (TPM) for granzymes, perforin and effector cytokines. Bar chart of mRNA expression (TPM) of (D) Granzymes A and B (E) Glucose transporters Slc2a1 and Slc2a3. (F-G) Fold change of PimdKO/WT protein (copies) from proteomics analysis described in Fig 4 vs mRNA (TPM) (F) highlighting in red proteins that are differentially expressed (FC>1.5, q<0.05) where mRNA is not substantially different (FC<1.2) and (G) highlighting in red protein and mRNA that are both differentially expressed (FC>1.5, q<0.05). (H) Estimated cytosolic ribosome content per cell (left), % ribosome of total cellular protein content (right). (I) Estimated protein copy number per cell of translation repressor PDCD4. (J) PDCD4 expression measured by flow cytometry on day 3 and 6 in IL-2 expanded WT vs Pim dKO CD8 T cells. (K) Estimated protein copy number per cell of EIF4A1. (L) adjusted ratio of PDCD4 : EIF4A1 (assuming 1 PDCD4 binds 2x EIF4A1) in WT and Pim dKO proteomes. (M) Protein synthesis measured by OPP incorporation in day 6 IL-2-expanded WT CTL treated for 24 hours with pan PIM kinase inhibitors PIM447 (5 µM) or AZD1208 (10 µM). 30 min cycloheximide (100 µg/mL) treatment gives no protein synthesis background control. Symbols in bar charts show biological replicates: error bars show mean ± S.D. Data are representative of (M) n=2 biological replicates collected over two independent experiments, (J) n = 2 biological replicates. Quantitative proteomics and RNAseq were performed on biological triplicates. * indicates q<0.05, fold-change (FC) shown on graph when q<0.05.

These observations could reflect that the PIM kinases are required for IL-2 to maximally control protein synthesis in CD8 T cells. To examine this hypothesis we first interrogated our proteomics and transcriptomics data. Consistent with this hypothesis we observed a small reduction in ribosome protein content in Pim dKO CTL, which scaled with the reduction in total protein content (Fig 5H). Moreover the translational repressor PDCD4 was increased in Pim dKO CTL (Fig 5I). This could be confirmed by flow cytometry (Fig 5J) and was underpinned at the mRNA level (Fig 5B). Similarly, 24 hour inhibition with pan PIM kinases inhibitors was sufficient to substantially increase PDCD4 protein expression (Fig S3F). We also noted that there was a small decrease in the expression of EIF4A1, a key component of the eIF4F translation initiation complex, in Pim dKO IL2 maintained CTL (Fig 5K). PDCD4 inhibits translation by binding EIF4A1 in a 2:1 ratio, preventing its interaction with the EIF4F complex (Suzuki et al., 2008). Together the changes in both proteins equated to an approximate halving of the ratio between EIF4A1 and PDCD4 (Fig 5L). These data all support that loss of PIM kinases could reduce protein synthesis in IL-2 maintained CTL. To test this hypothesis we used a single cell assay that quantifies the incorporation of an analogue of puromycin (OPP) into newly synthesized protein chains in the ribosome to assess cellular rates of protein synthesis. We found that 24 hours of treatment with pan-PIM kinase inhibitors PIM447 or AZD1208 caused a quantitative reduction in protein synthesis rates of IL-2 expanded CTL (Fig 5M). This data is consistent with PIM kinase effects on protein expression being via control of protein translation.

PIM1 and PIM2 regulate mTORc1 and sustain lymphoid homing in Pim1/Pim2-deficient effector CTL

The pattern of proteome remodelling in Pim dKO CTL had some phenotypic features in common with the previously reported changes that occur following inhibition of the Serine/Threonine kinase mTORc1 in CTL (Hukelmann et al., 2016; Howden et al., 2019). These included: increased expression of PDCD4 and decreased expression of cholesterol biosynthesis enzymes and effector molecules including granzymes, perforin, IFNγ and TNFα (Hukelmann et al., 2016; Howden et al., 2019). In this respect, mTORc1 activity in CTL is acutely sensitive to reduced availability of glucose (Rolf et al., 2013) and the reduced expression of glucose transporters in Pim dKO T cells might predict that these cells would have lower activity of mTORc1. We therefore assessed if PIM kinases were required for IL-2 activation of mTORc1. In these experiments we compared the phosphorylation of S6K1 (p70 S6-kinase 1) on its mTORC1 substrate sequence T389 in WT and Pim dKO CD8 T cells by western blot. These data show a modest reduction of mTORc1 activity in the Pim dKO T cells (Fig 6A). We also assessed the activity of S6K1 at a single cell level by quantifying the phosphorylation of its downstream target protein S6 ribosomal protein on S235/S236 by flow cytometry over the course of IL-2 driven CTL expansion. These data also show that Pim1/Pim2-deficiency reduced IL-2 induced mTORc1 dependent phosphorylation of S6 but did not compromise the ability of the Pim dKO cells to phosphorylate S6 in response to phorbol esters and calcium ionophores; pharmacological activators that induce RSK mediated S6 phosphorylation (Fig 6B).

PIM kinases regulate mTORc1 activity and lymphoid homing in effector CTL

(A) Western blot of pS6K T389 and pan S6K from day 6 WT and PimdKO IL-2 CTL (paired student t-test, *= p<0.05). (B) WT and Pim dKO T cells were mixed in a 50:50 ratio, activated with αCD3/αCD28 (both 0.5 µg/mL) and IL-2 (20 ng/mL) and expanded in IL-2 as per Fig 2B and pS6 (Ser235/236) measured after 1 hr +/-PdBU and ionomycin (positive control), +/-rapamycin (mTORc1 inhibitor, negative control) or no additional treatment. mRNA expression (TPM) from RNAseq analysis of IL-2 expanded WT and Pim dKO CTL described in Fig 5A for (C) cell homing receptors S1pr1, Ccr7 and Sell and (D) transcription factor Klf2. (E) WT and Pim dKO T cells were activated and expanded with IL-2 in separate cultures as per Fig 2B. On day 6 of culture WT and Pim dKO CTL were labelled with CFSE or CTV, mixed at a 50:50 ratio and transferred into C57Bl/6 recipient mice. Values indicate percentage of transferred cells detected in blood, lymph node or spleen 4 or 23 hours post-transfer that were WT or Pim dKO (one-way ANOVA). Symbols show biological replicates. Error bars show mean ± S.D. Data are representative of (A) n= 6, (B) n=2-4 collected across at least 2 independent experiments and (E) n= 6 recipient mice, from n=2 biological donor replicates. RNAseq was performed on biological triplicates. * q<0.05, ** q<0.01, *** q<0.001, fold-change (FC) shown on graph when q<0.05 but FC<1.5.

Previously, it has been shown that decreased glucose availability or inhibition of mTORc1 prevents the downregulation of CD62L that normally accompanies IL-2 induced CTL differentiation (Sinclair et al., 2008; Finlay et al., 2012). Thus, the reduced expression of glucose transporters and reduced activity of mTORc1 in Pim dKO IL-2 expanded CTL provides a potential explanation for the striking phenotype shown in Fig 3G, that IL-2 maintained Pim dKO CTL retain high levels of CD62L/L-selectin. The IL-2 CTL transcriptomics data showed there is increased expression of CD62L at the mRNA level in IL-2 maintained PimdKO CTL (Fig 5B, 6C). The expression of Sell (CD62L) mRNA is controlled by the transcription factor KLF2 and the expression of Klf2 mRNA is also increased in IL-2 maintained Pim dKO CTL (Fig 6D) as are mRNA levels for other KLF2 targets: the chemokine receptors Ccr7 and S1pr1 (Fig 6C).

CD62L/L-selectin controls T cell adhesion to the endothelium of high endothelial venules and is essential for lymphocyte transit from blood into secondary lymphoid tissue such as lymph nodes. CCR7, and the S1P1 receptor also direct migration by controlling T cell entry and egress respectively into secondary lymphoid tissue. In this context, the loss of CD62L, CCR7 and S1PR1 expression following immune activation reprograms the trafficking of effector T cells away from lymphoid tissue toward sites of inflammation in peripheral tissues (Sinclair et al., 2008; Nolz et al., 2011). Accordingly, the differences between Pim dKO and WT CTL in terms of CD62L expression could indicate that in addition to having reduced effector protein expression, Pim dKO CTL also do not switch their trafficking behaviour to that of a fully differentiated effector CTL and rather retain the capacity to home to lymphoid tissues. To test this possibility in vivo, we performed adoptive transfer experiments comparing the ability of WT effector CTL and Pim dKO CTL to home to secondary lymphoid tissues. In these experiments, αCD3/αCD28 activated WT or Pim dKO T cells were differentiated in IL-2 and cells then were labelled with either CFSE or CTV dyes; mixed at a ratio of 1:1, and adoptively transferred into C57BL/6 hosts. Fig 6E shows that Pim dKO CTL, but not WT CTL, retained substantial capacity to home to secondary lymphoid organs and, hence, accumulated in lymph nodes and spleen: PIM kinases are thus required for the normal reprogramming of CTL trafficking.

Discussion

The objective of the present study was to use quantitative analysis of T cell proteomes and transcriptomes to explore how deletion of the Serine/Threonine kinases PIM1 and PIM2 impacts CD8 T cell activation and differentiation. The expression of PIM1 and PIM2 is induced during immune activation of CD8 T cells by antigen and co-stimulation and then sustained to varying degrees by the cytokines IL2 or IL-15. The loss of PIM1 and PIM2 was shown to have no effect on the CD8 T cell programs controlled by antigen receptor/co-stimulation and had very little impact on the IL-15-driven transcriptional and proteomic programs in memory differentiated CD8 T cells. However, PIM1 and PIM2 were required for IL-2 programming of effector CD8 T cell differentiation. IL-2 controls essential cytolytic effector and metabolic programs in CTL, is a key cytokine for CD8 T cell anti-viral immunity and is also used to produce cytotoxic T cells for anti-cancer immunotherapy (Kalia and Sarkar, 2018). The understanding that PIM kinases mediate IL-2 control of the expression of glucose transporters, lipid metabolic enzymes and key cytolytic proteins such as granzymes A and B in effector CD8 T cells is thus fundamental information about how this key cytokine controls T cells. It was also important to understand that PIM kinases did not control the transcriptional programs that direct expression of glucose transporters and effector enzymes and the hence the full consequences of PIM kinase deletion could only be assessed by in depth analysis of the proteomes of Pim1/Pim2-deficient CTL and not from mRNA sequencing. The loss of PIM kinase activity reduced protein synthesis rates in CTL which explains the discordance between mRNA and protein reported herein. The reduction in protein synthesis was modest but previous studies have shown that modest reductions in protein synthesis and loss of mTORc1 activity impact CD8 T cell differentiation programs (Cornish et al., 2006; Hukelmann et al., 2016; Howden et al., 2019). PIM kinase regulation of protein translation has been reported previously in mouse B cell lymphoma and mouse embryonic fibroblasts (Schatz et al., 2011) and some of the key examples of proteins we found to be unaltered at the mRNA level but down-regulated at the protein level have prior reports of sensitivity to translational regulation in immune cells (Fehniger et al., 2007; Ricciardi et al., 2018). Reduced protein synthesis capacity in cells lacking PIM kinases could be caused by reduced mTORc1 activity combined with increased expression at the mRNA and protein level of the translational repressor PDCD4. Moreover, protein synthesis is highly energetically demanding and once CTL have reduced glucose transporter expression this would reduce glucose availability to fuel protein production. These data collectively point towards modulation of protein translation as an important target of PIM kinase activity downstream of IL-2 in CD8 T cells.

One other important insight from the current work is that PIM kinases are required for IL2 to fully repress expression of the adhesion molecule CD62L and the chemokine receptors CCR7 and S1PR1. These molecules play a key role in controlling T cell homing to secondary lymphoid tissue and are expressed at high levels in naïve and memory T cells. IL-2 induces down regulation of CD62L, CCR7 and S1PR1 expression as part of a program that causes effector cells to lose their capacity to home to secondary lymphoid tissues (Sinclair et al., 2008; Nolz et al., 2011). The present data show that PIM kinases can mediate this IL-2 repression of lymph node homing receptors and the ability of IL-2 to switch off the naïve T cell trafficking program. IL2 maintained Pim dKO CD8 effector T cells thus retain the capacity to home to secondary lymphoid tissues in vivo. This lymph-node homing pattern is also a memory CD8 T cell phenotype but here it is pertinent that the current in-depth analysis of the transcriptomes and proteomes of Pim1/Pim2-deficient IL-2 differentiated CD8 T cells show that the loss of PIM kinases does not cause effector T cells to differentiate instead into memory cells. For example, IL-2 maintained Pim dKO CD8 T cells do not re-express key molecules that maintain the stem-like properties of memory cells such as the transcription factor TCF7 and IL-7 receptor (Table S2, S6). Nor do Pim dKO CTL cultured in IL-2 have the mitochondrial proteome of a memory T cell. Hence Pim1/Pim2-deficiency prevents optimal differentiation of effector CTL but does not cause these cells to fully switch to a memory phenotype. This is worth knowing as pan-PIM kinase inhibitors are in development as anti-cancer drugs and it has been suggested that PIM inhibitors might be used as a strategy to generate more stem-memory-like CD8 T cells in combination with cancer immunotherapy treatments (Chatterjee et al., 2019; Clements and Warfel, 2022). Information about how the loss of PIM kinases will impact CD8 T cells is very relevant to informed usage of such drugs.

Finally, the selectivity of PIM kinases for IL-2 control of conventional CD8 T cell differentiation with no evidence that PIMs were important for IL-15 responses beyond a modest effect on cell proliferation was interesting. This failure to see a role for PIM signalling in conventional peripheral CD8 T cell differentiation is in contrast to observations that PIM1 and PIM2 direct IL15 control of the metabolic programming of intestinal IEL (James et al., 2021). A notable difference between these scenarios was that in IEL, IL15 was being used individually to drive immune activation and was sufficient to strongly induce PIM1 and PIM2 expression. In contrast, in our conventional CD8 T cell system we uncovered that that IL-15 does not sustain high levels of PIM kinases, which could possibly explain this discrepancy. However, high levels of PIM expression alone are not inevitably predictive of their importance. In this study we observed very high levels of both PIM1 and PIM2 in antigen activated CD8 T cells yet loss of PIM kinases there had no functional impact. There are thus clearly redundant kinase pathways that can substitute for the PIM kinases in the immediate response of CD8 T cells to immune activation. Previously it was suggested that mTORc1 may be one of these pathways (Fox et al., 2005), however we found no evidence for that here, and in fact observed that PIM kinase deficiency was upstream of a reduction in mTORc1 activity in IL2 expanded CTL. Thus, this examination of the role of PIM kinases in multiple signalling situations reveal cell context dependent regulatory roles for PIM kinases rather than hard wired functions that apply to every stimulatory situation in T cells.

Supporting information

eLife Assessment

This is an important study, as PIM1/2 control of protein synthesis in differentiated cells has implications beyond T cells. The evidence is convincing in that it makes extensive use of the mouse knockout model and validation in mouse T cells with inhibitors. A rescue experiment in mouse KO T cells would be even stronger than the inhibitor studies to validate the KO phenotype and the evidence would be truly impressive if the results from the rescue experiment support the working model. Extending the observations to human T cells would also be a step towards translation and would further increase the potential impact of the work.

Reviewer #1 (Public Review):

Summary and Strengths:

The study focuses on PIM1 and 2 in CD8 T cell activation and differentiation. These two serine/threonine kinases belong to a large network of Serine/Threonine kinases that acts following engagement of the TCR and of cytokine receptors and phosphorylates proteins that control transcriptional, translational and metabolic programs that result in effector and memory T cell differentiation. The expression of PIM1 and PIM2 is induced by the T-cell receptor and several cytokine receptors. The present study capitalized on high-resolution quantitative analysis of the proteomes and transcriptomes of Pim1/Pim2-deficient CD8 T cells to decipher how the PIM1/2 kinases control TCR-driven activation and IL-2/IL-15-driven proliferation, and differentiation into effector T cells.

Quantitative mass spectrometry-based proteomics analysis of naïve OT1 CD8 T cell stimulated with their cognate peptide showed that the PIM1 protein was induced within 3 hours of TCR engagement and its expression was sustained at least up to 24 hours. The kinetics of PIM2 expression was protracted as compared to that of PIM1. Such TCR-dependent expression of PIM1/2 correlated with the analysis of both Pim1 and Pim2 mRNA. In contrast, Pim3 mRNA was only expressed at very low levels and the PIM3 protein was not detected by mass spectrometry. Therefore, PIM1 and 2 are the major PIM kinases in recently activated T cells. Pim1/Pim2 double knockout (Pim dKO) mice were generated on a B6 background and found to express a lower number of splenocytes. No difference in TCR/CD28-driven proliferation was observed between WT and Pim dKO T cells over 3 days in culture. Quantitative proteomics of >7000 proteins further revealed no substantial quantitative or qualitative differences in protein content or proteome composition. Therefore, other signaling pathways can compensate for the lack of PIM kinases downstream of TCR activation.

Considering that PIM1 and PIM2 kinase expression is regulated by IL-2 and IL-15, antigen-primed CD8 T cells were expanded in IL-15 to generate memory phenotype CD8 T cells or expanded in IL-2 to generate effector cytotoxic T lymphocytes (CTL). Analysis of the survival, proliferation, proteome, and transcriptome of Pim dKO CD8 T cells kept for 6 days in IL-15 showed that PIM1 and PIM2 are dispensable to drive the IL-15-mediated metabolic or differentiation programs of antigen-primed CD8 T cells. Moreover, Pim1/Pim2-deficiency had no impact on the ability of IL-2 to maintain CD8 T cell viability and proliferation. However, WT CTL downregulated the expression of CD62L whereas the Pim dKO CTL sustained higher CD62L expression. Pim dKO CTL was also smaller and less granular than WT CTL. Comparison of the proteome of day 6 IL-2 cultured WT and Pim dKO CTL showed that the latter expressed lower levels of the glucose transporters, SLC2A1 and SLC2A3, of a number of proteins involved in fatty acid and cholesterol biosynthesis, and CTL effector proteins such as granzymes, perforin, IFNg, and TNFa. Parallel transcriptomics analysis showed that the reduced expression of perforin and some granzymes correlated with a decrease in their mRNA whereas the decreased protein levels of granzymes B and A, and the glucose transporters SLC2A1 and SLC2A3 did not correspond with decreased mRNA expression. Therefore, PIM kinases are likely required for IL-2 to maximally control protein synthesis in CD8 CTL. Along that line, the translational repressor PDCD4 was increased in Pim dKO CTL and pan-PIM kinase inhibitors caused a reduction in protein synthesis rates in IL-2-expanded CTL. Finally, the differences between Pim dKO and WT CTL in terms of CD62L expression resulted in Pim dKO CTL but not WT CTL retained the capacity to home to secondary lymphoid organs. In conclusion, this thorough and solid study showed that the PIM1/2 kinases shape the effector CD8 T cell proteomes rather than transcriptomes and are important mediators of IL2-signalling and CD8 T cell trafficking.

Weaknesses:

None identified by this reviewer.

Reviewer #2 (Public Review):

Summary:

Using a suite of techniques (e.g., RNA seq, proteomics, and functional experiments ex vivo) this paper extensively focuses on the role of PIM1/2 kinases during CD8 T-cell activation and cytokine-driven (i.e., IL-2 or IL-15) differentiation. The authors' key finding is that PIM1/2 enhances protein synthesis in response to IL-2 stimulation, but not IL-15, in CD8+ T cells. Loss of PIM1/2 made T cells less 'effector-like', with lower granzyme and cytokine production, and a surface profile that maintained homing towards secondary lymphoid tissue. The cytokines the authors focus on are IL-15 and Il-2, which drive naïve CD8 T cells towards memory or effector states, respectively. Although PIM1/2 are upregulated in response to T-cell activation and cytokine stimulation (e.g., IL-15, and to a greater extent, IL-2), using T cells isolated from a global mouse genetic knockout background of PIM1/2, the authors find that PIM1/2 did not significantly influence T-cell activation, proliferation, or expression of anything in the proteome under anti-CD3/CD28 driven activation with/without cytokine (i.e., IL-15) stimulation ex vivo. This is perhaps somewhat surprising given PIM1/2 is upregulated, albeit to a small degree, in response to IL-15, and yet PIM1/2 did not seem to influence CD8+ T cell differentiation towards a memory state. Even more surprising is that IL-15 was previously shown to influence the metabolic programming of intestinal intraepithelial lymphocytes, suggesting cell-type specific effects from PIM kinases. What the authors went on to show, however, is that PIM1/2 KO altered CD8 T cell proteomes in response to IL-2. Using proteomics, they saw increased expression of homing receptors (i.e., L-selectin, CCR7), but reduced expression of metabolism-related proteins (e.g., GLUT1/3 & cholesterol biosynthesis) and effector-function related proteins (e.g., IFNy and granzymes). Rather neatly, by performing both RNA-seq and proteomics on the same IL-2 stimulated WT vs. PIM1/2 KO cells, the authors found that changes at the proteome level were not corroborated by differences in RNA uncovering that PIM1/2 predominantly influence protein synthesis/translation. Effectively, PIM1/2 knockout reduced the differentiation of CD8+ T cells towards an effector state. In vivo adoptive transfer experiments showed that PIM1/2KO cells homed better to secondary lymphoid tissue, presumably owing to their heightened L-selectin expression (although this was not directly examined).

Strengths:

Overall, I think the paper is scientifically good, and I have no major qualms with the paper. The paper as it stands is solid, and while the experimental aim of this paper was quite specific/niche, it is overall a nice addition to our understanding of how serine/threonine kinases impact T cell state, tissue homing, and functionality. Of note, they hint towards a more general finding that kinases may have distinct behaviour in different T-cell subtypes/states. I particularly liked their use of matched RNA-seq and proteomics to first suggest that PIM1/2 kinases may predominantly influence translation (then going on to verify this via their protein translation experiment - although I must add this was only done using PIM kinase inhibitors, not the PIM1/2KO cells). I also liked that they used small molecule inhibitors to acutely reduce PIM1/2 activity, which corroborated some of their mouse knockout findings - this experiment helps resolve any findings resulting from potential adaptation issues from the PIM1/2 global knockout in mice but also gives it a more translational link given the potential use of PIM kinase inhibitors in the clinic. The proteomics and RNA seq dataset may be of general use to the community, particularly for analysis of IL-15 or IL-2 stimulated CD8+ T cells.

Weaknesses:

It would be good to perform some experiments in human T cells too, given the ease of e.g., the small molecule inhibitor experiment. Would also be good for the authors to include a few experiments where PIM1/2 have been transduced back into the PIM1/2 KO T cells, to see if this reverts any differences observed in response to IL-2 - although the reviewer notes that the timeline for altering primary T cells via lentivirus/CRISPR may be on the cusp of being practical such that functional experiments can be performed on day 6 after first stimulating T cells. Other experiments could also look at how PIM1/2 KO influences the differentiation of T cell populations/states during ex vivo stimulation of PBMCs or in vivo infection models using (high-dimensional) flow cytometry (rather than using bulk proteomics/RNA seq which only provide an overview of all cell combined). Alongside this, performing a PCA of bulk RNA seq/proteomes or Untreated vs. IL-2 vs. IL-15 of WT and PIM1/2 knockout T cells would help cement their argument in the discussion about PIM1/2 knockout cells being distinct from a memory phenotype.

Author response:

The following is the authors’ response to the original reviews

We thank the reviewers for their careful reading of our manuscript and their considered feedback. Please see our detailed response to reviewer comments inset below.

In addition to requested modifications we have also uploaded the proteomics data from 2 of the experiments contained within the manuscript onto the Immunological Proteome Resource (ImmPRes) website: immpres.co.uk making the data available in an easy-to-use graphical format for interested readers to interrogate and explore. We have added the following text to the data availability section (lines 1085-1091) to indicate this:

“An easy-to-use graphical interface for examining protein copy number expression from the 24-hour TCR WT and Pim dKO CD4 and CD8 T cell proteomics and IL-2 and IL-15 expanded WT and Pim dKO CD8 T cell proteomics datasets is also available on the Immunological Proteome Resource website: immpres.co.uk (Brenes et al., 2023) under the Cell type(s) selection: “T cell specific” and Dataset selection: “Pim1/2 regulated TCR proteomes” and “Pim1/2 regulated IL2 or IL15 CD8 T cell proteomes”.”

As well as indicating in figure legends where proteomics datasets are first introduced in Figures 1, 2 and 4 with the text:

“An interactive version of the proteomics expression data is available for exploration on the Immunological Proteome Resource website: immpres.co.uk”

Public Reviews:

Reviewer #1 (Public Review):

Summary and Strengths:

The study focuses on PIM1 and 2 in CD8 T cell activation and differentiation. These two serine/threonine kinases belong to a large network of Serine/Threonine kinases that acts following engagement of the TCR and of cytokine receptors and phosphorylates proteins that control transcriptional, translational and metabolic programs that result in effector and memory T cell differentiation. The expression of PIM1 and PIM2 is induced by the T-cell receptor and several cytokine receptors. The present study capitalized on high-resolution quantitative analysis of the proteomes and transcriptomes of Pim1/Pim2-deficient CD8 T cells to decipher how the PIM1/2 kinases control TCRdriven activation and IL-2/IL-15-driven proliferation, and differentiation into effector T cells.

Quantitative mass spectrometry-based proteomics analysis of naïve OT1 CD8 T cell stimulated with their cognate peptide showed that the PIM1 protein was induced within 3 hours of TCR engagement, and its expression was sustained at least up to 24 hours. The kinetics of PIM2 expression was protracted as compared to that of PIM1. Such TCRdependent expression of PIM1/2 correlated with the analysis of both Pim1 and Pim2 mRNA. In contrast, Pim3 mRNA was only expressed at very low levels and the PIM3 protein was not detected by mass spectrometry. Therefore, PIM1 and 2 are the major PIM kinases in recently activated T cells. Pim1/Pim2 double knockout (Pim dKO) mice were generated on a B6 background and found to express a lower number of splenocytes. No difference in TCR/CD28-driven proliferation was observed between WT and Pim dKO T cells over 3 days in culture. Quantitative proteomics of >7000 proteins further revealed no substantial quantitative or qualitative differences in protein content or proteome composition. Therefore, other signaling pathways can compensate for the lack of PIM kinases downstream of TCR activation.

Considering that PIM1 and PIM2 kinase expression is regulated by IL-2 and IL-15, antigen-primed CD8 T cells were expanded in IL-15 to generate memory phenotype CD8 T cells or expanded in IL-2 to generate effector cytotoxic T lymphocytes (CTL). Analysis of the survival, proliferation, proteome, and transcriptome of Pim dKO CD8 T cells kept for 6 days in IL-15 showed that PIM1 and PIM2 are dispensable to drive the IL-15mediated metabolic or differentiation programs of antigen-primed CD8 T cells. Moreover, Pim1/Pim2-deficiency had no impact on the ability of IL-2 to maintain CD8 T cell viability and proliferation. However, WT CTL downregulated the expression of CD62L whereas the Pim dKO CTL sustained higher CD62L expression. Pim dKO CTL was also smaller and less granular than WT CTL. Comparison of the proteome of day 6 IL-2 cultured WT and Pim dKO CTL showed that the latter expressed lower levels of the glucose transporters, SLC2A1 and SLC2A3, of a number of proteins involved in fatty acid and cholesterol biosynthesis, and CTL effector proteins such as granzymes, perforin, IFNg, and TNFa. Parallel transcriptomics analysis showed that the reduced expression of perforin and some granzymes correlated with a decrease in their mRNA whereas the decreased protein levels of granzymes B and A, and the glucose transporters SLC2A1 and SLC2A3 did not correspond with decreased mRNA expression. Therefore, PIM kinases are likely required for IL-2 to maximally control protein synthesis in CD8 CTL. Along that line, the translational repressor PDCD4 was increased in Pim dKO CTL and pan-PIM kinase inhibitors caused a reduction in protein synthesis rates in IL-2expanded CTL. Finally, the differences between Pim dKO and WT CTL in terms of CD62L expression resulted in Pim dKO CTL but not WT CTL retained the capacity to home to secondary lymphoid organs. In conclusion, this thorough and solid study showed that the PIM1/2 kinases shape the effector CD8 T cell proteomes rather than transcriptomes and are important mediators of IL2-signalling and CD8 T cell trafficking.

Weaknesses:

None identified by this reviewer.

Reviewer #2 (Public Review):

Summary:

Using a suite of techniques (e.g., RNA seq, proteomics, and functional experiments ex vivo) this paper extensively focuses on the role of PIM1/2 kinases during CD8 T-cell activation and cytokine-driven (i.e., IL-2 or IL-15) differentiation. The authors' key finding is that PIM1/2 enhances protein synthesis in response to IL-2 stimulation, but not IL-15, in CD8+ T cells. Loss of PIM1/2 made T cells less 'effector-like', with lower granzyme and cytokine production, and a surface profile that maintained homing towards secondary lymphoid tissue. The cytokines the authors focus on are IL-15 and Il-2, which drive naïve CD8 T cells towards memory or effector states, respectively. Although PIM1/2 are upregulated in response to T-cell activation and cytokine stimulation (e.g., IL-15, and to a greater extent, IL-2), using T cells isolated from a global mouse genetic knockout background of PIM1/2, the authors find that PIM1/2 did not significantly influence T-cell activation, proliferation, or expression of anything in the proteome under anti-

CD3/CD28 driven activation with/without cytokine (i.e., IL-15) stimulation ex vivo. This is perhaps somewhat surprising given PIM1/2 is upregulated, albeit to a small degree, in response to IL-15, and yet PIM1/2 did not seem to influence CD8+ T cell differentiation towards a memory state. Even more surprising is that IL-15 was previously shown to influence the metabolic programming of intestinal intraepithelial lymphocytes, suggesting cell-type specific effects from PIM kinases. What the authors went on to show, however, is that PIM1/2 KO altered CD8 T cell proteomes in response to IL-2. Using proteomics, they saw increased expression of homing receptors (i.e., L-selectin, CCR7), but reduced expression of metabolism-related proteins (e.g., GLUT1/3 & cholesterol biosynthesis) and effector-function related proteins (e.g., IFNy and granzymes). Rather neatly, by performing both RNA-seq and proteomics on the same IL2 stimulated WT vs. PIM1/2 KO cells, the authors found that changes at the proteome level were not corroborated by differences in RNA uncovering that PIM1/2 predominantly influence protein synthesis/translation. Effectively, PIM1/2 knockout reduced the differentiation of CD8+ T cells towards an effector state. In vivo adoptive transfer experiments showed that PIM1/2KO cells homed better to secondary lymphoid tissue, presumably owing to their heightened L-selectin expression (although this was not directly examined).

Strengths:

Overall, I think the paper is scientifically good, and I have no major qualms with the paper. The paper as it stands is solid, and while the experimental aim of this paper was quite specific/niche, it is overall a nice addition to our understanding of how serine/threonine kinases impact T cell state, tissue homing, and functionality. Of note, they hint towards a more general finding that kinases may have distinct behaviour in different T-cell subtypes/states. I particularly liked their use of matched RNA-seq and proteomics to first suggest that PIM1/2 kinases may predominantly influence translation (then going on to verify this via their protein translation experiment - although I must add this was only done using PIM kinase inhibitors, not the PIM1/2KO cells). I also liked that they used small molecule inhibitors to acutely reduce PIM1/2 activity, which corroborated some of their mouse knockout findings - this experiment helps resolve any findings resulting from potential adaptation issues from the PIM1/2 global knockout in mice but also gives it a more translational link given the potential use of PIM kinase inhibitors in the clinic. The proteomics and RNA seq dataset may be of general use to the community, particularly for analysis of IL-15 or IL-2 stimulated CD8+ T cells.

We thank the reviewer for their comments supporting the robustness and usefulness of our data.

Weaknesses:

It would be good to perform some experiments in human T cells too, given the ease of e.g., the small molecule inhibitor experiment.

The suggestions to check PIM inhibitor effects in human T cell is a good one. We think an ideal experiment would be to use naïve cord blood derived CD4 and CD8 cells as a model to avoid the impact of variability in adult PBMC and to really look at what PIM kinases do as T cells first respond to antigen and cytokines. In this context there is good evidence that the signalling pathways used by antigen receptors or the cytokines IL-2 and IL-15 are not substantially different in mouse and human. We have also previously compared proteomes of mouse and human IL-2 expanded cytotoxic T cells and they are remarkably similar. As such we feel that mature mouse CD8 T cells are a genetically tractable model to use to probe the signalling pathways that control cytotoxic T cell function. To repeat the full set of experiments observed within this study with human T cells would represent 1-year of work by an experienced postdoctoral fellow.

Unfortunately, the funding for the project has come to an end and there is no capacity to complete this work.

Would also be good for the authors to include a few experiments where PIM1/2 have been transduced back into the PIM1/2 KO T cells, to see if this reverts any differences observed in response to IL-2 - although the reviewer notes that the timeline for altering primary T cells via lentivirus/CRISPR may be on the cusp of being practical such that functional experiments can be performed on day 6 after first stimulating T cells.

A rescue experiment could indeed be informative, though of course comes with challenges/caveats with re-expressing both proteins that have been deleted at once and ability to control the level of PIM kinase that is re-expressed. This work using the Pim dKO mice was performed from 2019-2021 and was seriously impacted by the work restrictions during the COVID19 pandemic. We had to curtail all mouse colonies to allow animal staff to work within the legal guidelines. We had to make choices and the Pim1/2 dKO colony was stopped because we felt we had generated very useful data from the work but could not justify continued maintenance of the colony at such a difficult time. As such we no longer have this mouse line to perform these rescue experiments.

We have however, performed a limited number of retroviral overexpression studies in WT IL-2-expanded CTL, where T cells were transfected after 24 hours activation and phenotype measured on day 6 of culture. We chose to leave these out of the initial manuscript as these were overexpression under conditions where PIM expression was already high, rather than a true test of the ability of PIM1 or PIM2 to rescue the Pim dKO phenotype. A more robust test would also have required doing these overexpression experiments in IL-15 expanded or cytokine deprived CTL where PIM kinase expression is low, however, we ran out of time and funding to complete this work.

We have provided Author response image 1 below from the experiments performed in the IL-2 CTL for interested readers. The limited experiments that were performed do support some key phenotypes observed with the Pim dKO mice or PIM inhibitors, finding that PIM1 or PIM2 overexpression was sufficient to increase S6 phosphorylation, and provided a small further increase in GzmB expression above the already very high levels in IL-2 expanded CTL.

PIM1 or PIM2 overexpression drives increased GzmB expression and S6 phosphorylation in WT IL-2 CTL.

OT1 lymph node cell suspensions were activated for 24 hours with SIINFEKL peptide (10 ng/mL), IL-2 (20 ng/mL) and IL-12 (2 ng/mL) then transfected with retroviruses to drive expression of PIM1-GFP, PIM2-GFP fusion proteins or a GFP only control. T cells were split into fresh media and IL-2 daily and (A) GzmB expression and (B) S6 phosphorylation assessed by flow cytometry in GFP+ve vs GFP-ve CD8 T cells 5 days post-transfection (i.e. day 6 of culture). Histograms are representative of 2 independent experiments.

Other experiments could also look at how PIM1/2 KO influences the differentiation of T cell populations/states during ex vivo stimulation of PBMCs or in vivo infection models using (high-dimensional) flow cytometry (rather than using bulk proteomics/RNA seq which only provide an overview of all cells combined).

We did consider the idea of in vivo experiments with the Pim1/2 dKO mice but rejected this idea as the mice have lost PIM kinases in all tissues and so we would not be able to understand if any phenotype was CD8 T cell selective. To note the Pim1/2 dKO mice are smaller than normal wild type mice (discussed further below) and clearly have complex phenotypes. An ideal experiment would be to make mice with floxed Pim1 and Pim2 alleles so that one could use cre recombinase to make a T cell-specific deletion and then study the impact of this in in vivo models. We did not have the budget or ethical approval to make these mice. Moreover, this study was carried out during the COVID pandemic when all animal experiments in the UK were severely restricted. So our objective was to get a molecular understanding of the consequences of losing theses kinases for CD8 T cells focusing on using controlled in vitro systems. We felt that this would generate important data that would guide any subsequent experiments by other groups interested in these enzymes.

We do accept the comment about bulk population proteomics. Unfortunately, single cell proteomics is still not an option at this point in time. High resolution multidimensional flow cytometry is a valuable technique but is limited to looking at only a few proteins for which good antibodies exist compared to the data one gets with high resolution proteomics.

Alongside this, performing a PCA of bulk RNA seq/proteomes or Untreated vs. IL-2 vs. IL-15 of WT and PIM1/2 knockout T cells would help cement their argument in the discussion about PIM1/2 knockout cells being distinct from a memory phenotype.

We thank the reviewer for this very good suggestion. We have now included PCAs for the RNAseq and proteomics datasets of IL-2 and IL-15 expanded WT vs Pim dKO CTL in Fig S5 and added the following text to the discussion section of the manuscript (lines 429-431):

“… and PCA plots of IL-15 and IL-2 proteomics and RNAseq data show that Pim dKO IL-2 expanded CTL are still much more similar to IL-2 expanded WT CTL than to IL-15 expanded CTL (Fig S5)”.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

In panel B of Figure S1, are the smaller numbers of splenocytes found in dKO fully accounted for by a reduction in the numbers of T cells or also correspond to a reduction in B cell numbers? Are the thymus and lymph nodes showing the same trend?

We’re happy to clarify on this.

Since we were focused on T cell phenotypes in the paper this is what we have plotted in this figure, however there is also a reduction in total number of B, NK and NKT cells in the Pim dKO mice (see James et al, Nat Commun, 2021 for additional subset percentages). We find that all immune subsets we have measured make up the same % of the spleen in Pim dKO vs WT mice (we show this for T cell subsets in what was formerly Fig S1C and is now Fig S1A), the total splenocyte count is just lower in the Pim dKO mice (which we show in what was formerly Fig S1B and is now Fig S1C). To note, the Pim dKO mice were smaller than their WT counterparts (though we have not formally weighed and quantified this) and we think this is likely the major factor leading to lower total splenocyte numbers.

We have not checked the thymus so can’t comment on this. We can confirm that lymph nodes from Pim dKO mice had the same number and % CD4 and CD8 T cells as in WT.

For our in vitro studies we have made sure to either use co-cultures or for single WT and Pim dKO cultures to equalise starting cell densities between wells to account for the difference in total splenocyte number. We have now clarified this point in the methods section lines 682-684

“For generation of memory-like or effector cytotoxic T lymphocytes (CTL) from mice with polyclonal T cell repertoires, LN or spleen single cell suspensions at an equal density for WT and Pim dKO cultures (~1-3 million live cells/mL)….”

Reviewer #2 (Recommendations For The Authors):

Line 89-99 - PIM kinase expression is elevated in T cells in autoimmunity and inhibiting therefore may make some sense if PIM is enhancing T cell activity. Why then would you use an inhibitor in cancer settings? This needs better clarification for readers, with reference to T cells, particularly given this is an important justification for looking at PIM kinases in T cells.

We thank the reviewer for highlighting the lack of clarity in our explanation here.

PIM kinase inhibitors alone are proposed as anti-tumour therapies for select cancers to block tumour growth. However so far these monotherapies haven’t been very effective in clinical trials and combination treatment options with a number of strategies are being explored. There are two lines of logic for why PIM kinase inhibitors might be a good combination with an e.g. anti-PD1 or adoptive T cell immunotherapy. 1) PIM kinase inhibition has been shown to reduce inhibitory/suppressive surface proteins (e.g. PDL1) and cytokine (e.g. TGFbeta) expression in tumour cells and macrophages in the tumour microenvironment. 2) Inhibiting glycolysis and increasing memory/stem-like phenotype has been identified as desirable for longer-lasting more potent anti-tumour T cell immunity. PIM kinase inhibition has been shown to reduce glycolytic function and increase several ‘stemness’ promoting transcription factors e.g. TCF7 in a previous study. Controlled murine cancer models have shown improvement in clearance with the combination of pan-Pim kinase inhibitors and anti-PD1/PDL1 treatments (Xin et al, Cancer Immunol Res, 2021 and Chatterjee et al, Clin Cancer Res 2019).

It is worth noting, this is seemingly contradictory with other studies of Pim kinases in T cells that have generally found Pim1/2/3 deletion or inhibition in T cells to be suppressive of their function.

We have clarified this reasoning/seeming conflict of results in the introductory text as follows (lines 90-101):

“PIM kinase inhibitors have also entered clinical trials to treat some cancers (e.g. multiple myeloma, acute myeloid leukaemia, prostate cancer), and although they have not been effective as a monotherapy, there is interest in combining these with immunotherapies. This is due to studies showing PIM inhibition reducing expression of inhibitory molecules (e.g. PD-L1) on tumour cells and macrophages in the tumour microenvironment and a reported increase of stem-like properties in PIM-deficient T cells which could potentially drive longer lasting anti-cancer responses (Chatterjee et al., 2019; Xin et al., 2021; Clements and Warfel, 2022). However, PIM kinase inhibition has also generally been shown to be inhibitory for T cell activation, proliferation and effector activities (Fox et al., 2003; Mikkers et al., 2004; Jackson et al., 2021) and use of PIM kinase inhibitors could have the side effect of diminishing the anti-tumour T cell response.”

Line 93 - The use of 'some cancers' is rather vague and unscientific - please correct phrasing like this. The same goes for lines 54 and 77 (some kinases and some analyses).

We have clarified the sentence in what is now Line 91 to include examples of some of the cancers that PIM kinase inhibitors have been explored for (see text correction in response to previous reviewer comment), which are predominantly haematological malignancies. The use of the phrase ‘some kinases’ and ‘some analyses’ in what are now Lines 52 and 75 is in our view appropriate as the subsequent sentence/(s) provide specific details on the kinases and analyses that are being referred to.

Lines 146-147 - Could it be that rather than redundancies, PIM KO is simply not influential on TCR/CD28 signalling in general but influences other pathways in the T cell?

We agree that the lack of PIM1/2 effect could also be because PIM targets downstream of TCR/CD28 are not influential and have clarified the text as follows (lines 156-161):

“These experiments quantified expression of >7000 proteins but found no substantial quantitative or qualitative differences in protein content or proteome composition in activated WT versus Pim dKO CD4 and CD8 T cells (Fig 1G-H) (Table S1). Collectively these results indicate that PIM kinases do not play an important unique role in the signalling pathways used by the TCR and CD28 to control T cell activation.”

Line 169 - Instead of specifying control - maybe put upregulate or downregulate for clarity.

We have changed the text as per reviewer suggestion (see line 183)

Line 182-183 - I would move the call out for Figure 2D to after the last call out for Figure 2C to make it more coherent for readers.

We have changed the text as per reviewer suggestion (see lines 197-200)

Line 190 - 14,000 RNA? total, unique? mRNA?

These are predominantly mRNA since a polyA enrichment was performed as part of the standard TruSeq stranded mRNA sample preparation process, however, a small number of lncRNA etc were also detected in our RNA sequencing. We left the results in as part of the overall analysis since it may be of interest to others but don’t look into it further. We do mention the existence of the non-mRNA briefly in the subsequent sentence when discussing the total number of DE RNA that were classified as protein coding vs non-coding.

We have edited this sentence as follows to more accurately reflect that the RNA being referred to is polyA+ (lines 205-207):

“The RNAseq analysis quantified ~14,000 unique polyA+ mRNA and using a cut off of >1.5 fold-change and q-value <0.05 we saw that the abundance of 381 polyA+ RNA was modified by Pim1/Pim2-deficiency (Fig 2E) (Table S2A).

Questions/points regarding figures:

Figure 1 - Is PIM3 changed in expression with the knockout of PIM1/2 in mice? Although the RNA is low could there be some compensation here? The authors put a good amount of effort in to showing that mouse T cells do not exhibit differences from knocking out pim1/2 i.e., Efforts have been made to address this using activation markers and cell size, cytokines, and proliferation and proteomics of activated T cells. What do the resting T cells look like though? Although TCR signalling is not impacted, other pathways might be. Resting-state comparison may identify this.

In all experiments Pim3 mRNA was only detected at very low levels and no PIM3 protein was detected by mass spectrometry in either wild type or PIM1/2 double KO TCR activated or cytokine expanded CD8 T cells (See Tables S1, S3, S4). There was similarly no change in Pim3 mRNA expression in RNAseq of IL-2 or IL-15 expanded CD8 T cells (See Tables S2, S6). While we have not confirmed this in resting state cells for all the conditions examined, there is no evidence that PIM3 compensates for PIM1/2deficiency or that PIM3 is substantially expressed in T cells.

Figure 1A&B - Does PIM kinase stay elevated when removing TCR stimulus? During egress from lymph node and trafficking to infection/tumour/autoimmune site, T cells experience a period of 'rest' from T-cell activation so is PIM upregulation stabilized, or does it just coincide with activation? This could be a crucial control given the rest of the study focuses on day 6 after initial activation (which includes 4 days of 'rest' from TCR stimulation). Nice resolution on early time course though.

This is an interesting question. Unfortunately, we do not know how sensitive PIM kinases are to TCR stimulus withdrawal, as we have not tried removing the TCR stimulus during early activation and measuring PIM expression.

Based on the data in Fig 2A there is a hint that 4 hours withdrawal of peptide stimulus may be enough to lose PIM1/2 expression (after ~36 hrs of TCR activation), however, we did not include a control condition where peptide is retained within the culture. Therefore, we cannot resolve this question from the current experimental data, as this difference could also be due to a further increase in PIMs in the cytokine treated conditions rather than a reduction in expression in the no cytokine condition. This ~36-hour time point is also at a stage where T cells have become more dependent on cytokines for their sustained signalling compared to TCR stimulus.

It is worth noting that PIM kinases are thought to have fairly short mRNA and protein half lives (~5-20 min for PIM1 in primary cells, ~10 min – 1 hr for PIM2). This is consistent with previous observations that cytotoxic T cells need sustained IL-2/Jak signalling to sustain PIM kinase expression, e.g. in Rollings et al (2018) Sci Signaling, DOI:10.1126/scisignal.aap8112 . We would therefore expect that sustained signalling from some external signalling receptor whether this is TCR, costimulatory receptors or cytokines is required to drive Pim1/2 mRNA and protein expression.

Figure 1D - the CD4 WT and Pim dKO plots are identical - presumably a copying error - please correct.

We apologise for the copying error and have amended the manuscript to show the correct data. We thank the reviewer for noticing this mistake.

In Figure 1H - there is one protein found significant - would be nice to mention what this is - for example, if this is a protein that influences TCR levels this could be quite important.

The protein is Phosphoribosyl Pyrophosphate synthase 1 like 1 (Prps1l1).

This was a low confidence quantification (based on only 2 peptides) with no known function in T cells. Based on what is known, this gene is predominantly expressed in the testis (though also detected in spleen, lung, liver). A whole-body KO mouse found no difference in male fertility. No further phenotype has been reported in this mouse. See: Wang et al (2018) Mol Reprod Dev, DOI: 10.1002/mrd.23053

We have added the following text to the legend of Figure 1H to address this protein:

“Phosphoribosyl Pyrophosphate synthase 1 like 1 (Prps1l1), was found to be higher in Pim dKO CD8 T cells, but was a low confidence quantification (based on only 2 unique peptides) with no known function in T cells.”

Figure S1 - In your mouse model the reduction in CD4 T cells is quite dramatic in the spleen - is this reduced homing or reduced production of T cells through development?

Could you quantify the percentage of CD45+ cells that are T cells from blood too? Would be good to have a more thorough analysis of this new mouse model.

We apologise for the lack of clarity around the Pim dKO mouse phenotype. Something we didn’t mention previously due to a lack of a formal measurement is that the Pim dKO mice were typically smaller than their WT counterparts. This is likely the main reason for total splenocytes being lower in the Pim dKO mice - every organ is smaller. It is not a phenotype reported in Pim1/2 dKO mice on an FVB background, though has been reported in the Pim1/2/3 triple KO mouse before (see Mikkers et al, Mol Cell Biol 2004 doi: 10.1128/MCB.24.13.6104-6115.2004).

The % cell type composition of the spleen is equivalent between WT and Pim dKO mice and as mentioned above, was controlled for when setting up of our in vitro cultures.

We have revised the main text and changed the order of the panels in Fig S1 to make this caveat clearer as follows (lines 138-144):

“There were normal proportions of peripheral T cells in spleens of Pim dKO mice (Fig S1A) similar to what has been reported previously in Pim dKO mice on an FVB/N genetic background (Mikkers et al., 2004), though the total number of T cells and splenocytes was lower than in age/sex matched wild-type (WT) mouse spleens (Fig S1B-C). This was not attributable to any one cell type (Fig S1A)(James et al., 2021) but was instead likely the result of these mice being smaller in size, a phenotype that has previously been reported in Pim1/2/3 triple KO mice (Mikkers et al., 2004).”

Figure S1C - why are only 10-15% of the cells alive? Please refer to this experiment in the main text if you are going to include it in the supplementary figure.

With regards what was previously Fig S1C (now Fig S1A) we apologise for our confusing labelling. We were quoting these numbers as the percentage of live splenocytes (i.e. % of live cells). Typically ~80-90% of the total splenocytes were alive by the time we had processed, stained and analysed them by flow cytometry direct ex vivo. Of these CD4 and CD8 T cells made up ~%10-15 of the total live splenocytes (with most of the rest of the live cells being B cells).

We have modified the axis to say “% of splenocytes” to make it clearer that this is what we are plotting.

Figure S1 - Would be good to show that the T cells are truly deficient in PIM1/2 in your mice to be absolutely sure. You could just make a supplementary plot from your mass spec data.

This is a good suggestion and we have now included this data as supplementary figure 2.

To note, due to the Pim1 knockout mouse design this is not as simple as showing presence or absence of total PIM1 protein detection in this instance.

To elaborate: the Pim1/Pim2 whole body KO mice used in this study were originally made by Prof Anton Berns’ lab (Pim1 KO = Laird et al Nucleic Acids Res, 1993, doi: 10.1093/nar/21.20.4750, with more detail on deletion construct in te Riele, H. et al, Nature,1990, DOI: 10.1038/348649a0; Pim2 KO = Mikkers et al, Mol Cell Biol, 2004, DOI: 10.1128/MCB.24.13.6104-6115.2004). They were given to Prof Victor Tybulewicz on an FVB/N background. He then backcrossed them onto the C57BL/6 background for > 10 generations then gave them to us to intercross into Pim1/2 dKO mice on a C57BL/6 background.

The strategy for Pim1 deletion was as follows:

A neomycin cassette was recombined into the Pim1 gene in exon 4 deleting 296 Pim1 nucleotides. More specifically, the 98th pim-1 codon (counted from the ATG start site = the translational starting point for the 34 kDa isoform of PIM1) was fused in frame by two extra codons (Ser, Leu) to the 5th neo codon (pKM109-90 was used). The 3'-end of neo included a polyadenylation signal. The cassette also contains the PyF101 enhancer (from piiMo +PyF101) to ensure expression of neo on homologous recombination in ES cells.

Collectively this means that the PIM1 polypeptide is made prior to amino acid 98 of the 34 kDa isoform but not after this point. This deletes functional kinase activity in both the 34 kDa and 44 kDa PIM1 isoforms. Ablation of PIM1 kinase function using this KO was verified via kinase activity assay in Laird et al. Nucelic Acids Res 1993.

The strategy to delete Pim2 was as follows:

“For the Pim2 targeting construct, genomic BamHI fragments encompassing Pim2 exons 1, 2, and 3 were replaced with the hygromycin resistance gene (Pgp) controlled by the human PGK promoter.” (Mikkers et al Mol Cell Biol, 2004)

The DDA mass spectrometry data collected in Fig 1 G-H and supplementary table 1 confirmed we do not detect peptides from after amino acid residue 98 in PIM1 (though we do detect peptides prior to this deletion point) and we do not detect peptides from the PIM2 protein in the Pim dKO mice. Thus confirming that no catalytically active PIM1/PIM2 proteins were made in these mice.

We have added a supplementary figure S2 showing this and the following text (Lines 155-156):

“Proteomics analysis confirmed that no catalytically active PIM1 and PIM2 protein were made in Pim dKO mice (Fig S2).”

Figure 2A - I found the multiple arrows a little confusing - would just use arrows to indicate predicted MW of protein and stars to indicate non-specific. Why are there 3 bands/arrows for PIM2?

The arrows have now been removed. We now mention the PIM1 and PIM2 isoform sizes in the figure legend and have left the ladder markings on the blots to give an indication of protein sizes. There are 2 isoforms for PIM1 (34 and 44 kDa) in addition to the nonspecific band and 3 isoforms of PIM2 (40, 37, 34 kDa, though two of these isoform bands are fairly faint in this instance). These are all created via ribosome use of different translational start sites from a single Pim1 or Pim2 mRNA transcript.

The following text has been added to the legend of Fig 2A:

“Western blots of PIM1 (two isoforms of 44 and 34 kDa, non-specific band indicated by *), PIM2 (three isoforms of 40, 37 and 34 kDa) or pSTAT5 Y694 expression.”

Figure 2A - why are the bands so faint for PIM1/2 (almost non-existent for PIM2 under no cytokine stim) here yet the protein expression seems abundant in Figure 1B upon stim without cytokines? Is this a sensitivity issue with WB vs proteomics? My apologies if I have missed something in the methods but please explain this discrepancy if not.

There is differing sensitivity of western blotting versus proteomics, but this is not the reason for the discrepancy between the data in Fig 1B versus 2A. These differences reflect that Fig1 B and Fig 2A contrast PIM levels in two different sets of conditions and that while proteomics allows for an estimate of ‘absolute abundance’ Western blotting only shows relative expression between the conditions assessed.

To expand on this… Fig 1B proteomics looks at naïve versus 24 hr aCD3/aCD28 TCR activated T cells. The western blot data in Fig 2A looks at T cells activated for 1.5 days with SIINFEKL peptide and then washed free of the media containing the TCR stimulus and cultured with no stimulus for 4 or 24 hrs hours and contrast this with cells cultured with IL-2 or IL-15 for 4 or 24 hours. All Fig 2A can tell us is that cytokine stimuli increases and/or sustains PIM1 and PIM2 protein above the level seen in TCR activated cells which have not been cultured with cytokine for a given time period. Overexposure of the blot does reveal detectable PIM1 and PIM2 protein in the no cytokine condition after 4 hrs. Whether this is equivalent to the PIM level in the 24 hr TCR activated cells in Fig 1B is not resolvable from this experiment as we have not included a sample from a naïve or 24 hr TCR activated T cell to act as a point of reference.

Figure 4F - Your proteomics data shows substantial downregulation in proteomics data for granzymes and ifny- possibly from normalization to maximise the differences in the graph - and yet your flow suggests there are only modest differences. Can you explain why a discrepancy in proteomics and flow data - perhaps presenting in a more representative manner (e.g., protein counts)?

The heatmaps are a scaled for ‘row max’ to ‘row min’ copy number comparison on a linear scale and do indeed visually maximise differences in expression between conditions. This feature of these heatmaps is also what makes the lack of difference in GzmB and GzmA at the mRNA heatmap in Fig 5C quite notable.

We have now included bar graphs of Granzymes A and B and IFNg protein copy number in Figure 4 (see new Fig 4G-H) to make clearer the magnitude of the effect on the major effector proteins involved in CTL killing function. It is worth noting that flow cytometry histograms from what was formerly Fig 4G (now Fig 4I) are on a log-scale so the shift in fluorescence does generally correspond well with the ~1.7-2.75-fold reduction in protein expression observed.

Figure 4G - did you use isotype controls for this flow experiment? Would help convince labelling has worked - particularly for low levels of IFNy production.

We did not use isotype controls in these experiments but we are using a well validated interferon gamma antibody and very carefully colour panel/compensation controls to minimise background staining. The only ways to be 100% confident that an antibody is selective is to use an interferon gamma null T cell which we do not have. We do however know that the antibody we use gives flow cytometry data consistent with other orthogonal approaches to measure interferon gamma e.g. ELISA and mass spectrometry.

Figure 5M - why perform this with just the PIM kinase inhibitors? Can you do this readout for the WT vs. PIM1/2KO cells too? This would really support your claims for the paper about PIM influencing translation given the off-target effects of SMIs.

Regrettably we have not done this particular experiment with the Pim dKO T cells. As mentioned above, due to this work being performed predominantly during the COVID19 pandemic we ultimately had to make the difficult decision to cease colony maintenance. When work restrictions were lifted we could not ethically or economically justify resurrecting a mouse colony for what was effectively one experiment, which is why we chose to test this key biological question with small molecule inhibitors instead.

We appreciate that SMIs have off target effects and this is why we used multiple panPIM kinase inhibitors for our SMI validation experiments. While the use of 2 different inhibitors still doesn’t completely negate the concern about possible off-target effects, our conclusions re: PIM kinases and impact on proteins synthesis are not solely based on the inhibitor work but also based on the decreased protein content of the PIM1/2 dKO T cells in the IL-2 CTL, and the data quantifying reductions in levels of many proteins but not their coding mRNA in PIM1/2dKO T cells compared to controls.

Version 2

Abstract

Abstract

Integration of a large network of kinase signalling pathways co-ordinates changes in the transcription, translation and metabolic events required for T cell activation and differentiation. The present study explores the role of the Serine/Threonine kinases PIM1 and PIM2 in controlling murine CD8 T lymphocyte antigen receptor-mediated activation and differentiation in response to the cytokines Interleukin 2 (IL-2) or IL-15. We show that PIM kinases are dispensable for the differentiation programs controlled by the antigen-receptor and IL-15. There is however a selective role for the PIM kinases in the context of IL-2 regulation of CD8 T cell fate. One key insight was that the PIM kinases controlled the migratory capabilities of effector CD8 T cells, with Pim1/Pim2-deficient CD8 T cells unable to fully switch off the naïve T cell chemokine and adhesion receptor program during effector differentiation. PIM kinases were also needed for IL-2 to sustain high expression of the glucose transporters SLC2A1 and SLC2A3 and to maintain activity of the nutrient sensing kinase mTORc1. Strikingly, PIM kinases did not have a dominant impact on IL-2-driven transcriptional programs but rather selectively modulated protein synthesis to shape cytotoxic T cell proteomes. This study reveals a selective role of PIM kinases in IL-2 control of CD8 T cells and highlights how regulated changes in protein synthesis can impact T cell phenotypes.

Introduction

CD8 T lymphocytes are key effector cells for adaptive immune responses to viruses, intracellular pathogens and tumours. Their clonal expansion and differentiation to cytotoxic effector or memory CD8 T cells is controlled by signalling pathways initiated by the T cell antigen receptor (TCR), co-stimulatory molecules and cytokines. Receptor engagement by immune stimuli triggers the activation of cytosolic tyrosine kinases that then direct the activity of a diverse network of Serine/Threonine kinases, disseminating signals from plasma membrane to cell interior. The Src family and ZAP70 Tyrosine kinases initiate this process downstream of the TCR, with the Janus kinases (JAK) as the main signal transducer for important cytokine regulators of T cell differentiation, IL-2 and IL-15 (Cantrell, 2015; Saravia et al., 2020). The subsequent network of Serine/Threonine kinases that is engaged beyond these initiating events coordinates phosphorylation of a diverse array of protein targets.

These protein targets include chromatin regulators, transcription factors, translation machinery, cytoskeletal proteins and metabolic enzymes, which then control the changes in transcriptional, translational and metabolic programs that drive effector and memory T cell differentiation (Navarro et al., 2011; Tan et al., 2017; Locard-Paulet et al., 2020).

CD8 T cells express ∼200 Serine/Threonine protein kinases (Navarro et al., 2014; Howden et al., 2019). There is a good understanding of the roles of some Serine/Threonine kinases in CD8 T cells, including the role of the diacylglycerol regulated kinases, MAP kinases ERK1 and 2, Phosphatidylinositol (3,4,5) triphosphate-controlled kinases PDK1 and AKT, and the nutrient sensing kinases mammalian Target of Rapamycin complex 1 (mTORc1) and AMP- activated protein kinase (AMPK) (Finlay et al., 2012; Rolf et al., 2013; Blagih et al., 2015; Cantrell, 2015; Hukelmann et al., 2016; Tan et al., 2017; Howden et al., 2019; Saravia et al., 2020; Damasio et al., 2021; Spinelli et al., 2021). However, beyond these core kinases there is much still to learn about how many of the ∼200 Serine/Threonine kinases expressed in CD8 T cells contribute to the control of T cell function. In this context, the PIM kinases are a family of Serine/Threonine kinases that are strongly and rapidly induced in T cells responding to immune stimuli (Aho et al., 2005; Fox et al., 2005; Peperzak et al., 2010; Jackson et al., 2012). The PIM family contains 3 Serine/Threonine kinases, PIM1, PIM2 and PIM3, of which T lymphocytes predominantly express PIM1 and PIM2 (Mikkers et al., 2004; Fox et al., 2005). As PIM kinases are constitutively active, their participation in signalling cascades is dictated by regulated changes in their protein expression levels (Warfel and Kraft, 2015). PIM1 and PIM2 have little to no expression in naïve T cells but are induced by T-cell receptor activation and costimulation (e.g. CD27) (Wingett et al., 1996; Fox et al., 2005; Peperzak et al., 2010). PIM kinases are also direct transcriptional targets of JAK/STAT signalling pathways and as such PIM1 and/or PIM2 are rapidly induced in T cells downstream of a diverse array of cytokines, including multiple common cytokine receptor γ-chain family cytokines, IL-6 and IL-12 (Matikainen et al., 1999; Aho et al., 2005; Fox et al., 2005; Warfel and Kraft, 2015; Buchacher et al., 2023).

There have been some analyses of how deletion of PIM kinases impacts CD8 T cell function. A mild reduction in Pim1/Pim2/Pim3-deficient CD4 and CD8 T cell proliferation has been observed in response to sub-optimal TCR stimulation and IL-2 (Mikkers et al., 2004) and there is evidence that PIM1 and PIM2 control survival pathways in naïve CD4 and CD8 T cells in concert with mTORc1 (Fox et al., 2005). Pim1/Pim2/Pim3 deletion is reported to promote expression of stemness and memory-associated transcription factors including Bcl6 and Tcf7 (Chatterjee et al., 2019) and PIM kinase activity has been linked with maintenance of memory CD8 T cell survival (Knudson et al., 2017). It has also been shown recently that Pim1/Pim2 deficiency prevented mTORc1 activation and the proliferation of non- conventional intra-epithelial gut CD8 T lymphocytes (IEL) in response to the cytokine IL-15 (James et al., 2021).

Elevated PIM kinase expression has been noted in patient T cells for a number of autoimmune conditions, including rheumatoid arthritis, uveitis and coeliac disease, and PIM kinase inhibition has been proposed as a possible intervention to dampen autoimmunity (James et al., 2021; Maney et al., 2021; Li et al., 2022). PIM kinase inhibitors have also entered clinical trials to treat some cancers (e.g. multiple myeloma, acute myeloid leukaemia, prostate cancer), and although they have not been effective as a monotherapy, there is interest in combining these with immunotherapies. This is due to studies showing PIM inhibition reducing expression of inhibitory molecules (e.g. PD-L1) on tumour cells and macrophages in the tumour microenvironment and a reported increase of stem-like properties in PIM-deficient T cells which could potentially drive longer lasting anti-cancer responses (Chatterjee et al., 2019; Xin et al., 2021; Clements and Warfel, 2022). However, PIM kinase inhibition has also generally been shown to be inhibitory for T cell activation, proliferation and effector activities (Fox et al., 2003; Mikkers et al., 2004; Jackson et al., 2012) and use of PIM kinase inhibitors could have the side effect of diminishing the anti- tumour T cell response. Given the role of CD8 T cells in anti-cancer immunity, it is important to understand how these kinases control CD8 T cell function. However, there has not yet been a systematic analysis of the role of PIM kinases in conventional CD8 T cells as they activate and differentiate to form effector and memory cells. Accordingly, the current study, uses high resolution quantitative analysis of Pim1/Pim2-deficient CD8 T cell proteomes and transcriptomes to obtain a comprehensive and unbiased understanding of how PIM kinases control CD8 T cell TCR-driven activation, and IL-2 or IL-15-driven proliferation, and differentiation. The study reveals a selective role for PIM kinases as mediators of IL-2 signalling to control CD8 T cell trafficking and effector functionality. We uncover that the dominant impact of PIM kinase loss was to shape effector CD8 T cell proteomes rather than transcriptomes, reflecting PIM kinase modulation of protein synthesis.

Results

PIM1 and PIM2 are strongly induced by antigen receptor engagement but are dispensable for T cell activation

T cell antigen receptor (TCR) -driven activation of naïve T cells initiates their clonal expansion and differentiation into effector and memory cells. Quantitative mass spectrometry-based proteomics analysis of naïve OT1 CD8 T cell stimulated with their cognate peptide over time (Marchingo et al., 2020) showed that PIM1 protein was strongly induced within 3 hours of TCR engagement, which was sustained at a 24 hour time point (Fig 1A). PIM2 expression was not detected until 9 hours post-stimulation but was also sustained at the 24-hour time point. 24-hour polyclonal activation of CD8 T cells with αCD3/αCD28 agonist antibodies similarly induced expression of PIM1 and PIM2 proteins (Fig 1B, Table S1). These increases in PIM kinases expression at a protein level were underpinned by transcriptional reprograming, with both Pim1 and Pim2 mRNA very lowly expressed in naïve CD8 T cells and strongly upregulated by immune activation (Fig 1C) (Spinelli et al., 2021). Consistent with previous reports of Pim1 and Pim2 being the major PIM kinases in T cells (Mikkers et al., 2004; Fox et al., 2005), Pim3 mRNA was only expressed at very low levels in naïve and activated CD8 T cells (Fig 1C), with PIM3 protein not detected at all by mass spectrometry (Table S1).

Pim1 and Pim2 are strongly TCR-induced, but dispensable for T cell activation

Estimated copies per cell of PIM1 and PIM2 protein from quantitative proteomics analysis of (A) OT1 CD8 T cells stimulated with SIINFEKL peptide for indicated times from published dataset (Marchingo et al., 2020) or (B) naïve ex vivo and 24 hour αCD3/αCD28 (TCR) activated WT CD8 T cells (see Fig 1G-H) for further details. (C) Fragments per kilobase million (FPKM) of Pim1, Pim2 and Pim3 mRNA from published (Spinelli et al., 2021) bulk RNAseq analysis of naïve and 24 hour gp33-41 peptide stimulated P14 CD8 T cells. Lymph node cell suspensions from C57BL/6 (WT) and Pim1KO/Pim2KO (Pim dKO) mice were activated for 24 hours with αCD3/αCD28 (both 0.5 µg/mL) and CD4 and CD8 T cell (D) FSC-A SSC-A profiles, (E) expression of surface activation markers (CD25, CD44, CD71) or CD8 T cell intracellular IFNγ were measured by flow cytometry. (F) Lymph node single cell suspensions from WT and Pim dKO mice were labelled with CellTrace Violet (CTV), activated with αCD3/αCD28 (both 0.5 µg/mL) and CD4 and CD8 T cell CTV proliferation profiles were measured at indicated time points. (G-H) Lymph node cell suspensions from WT and Pim dKO mice were stimulated for 24 hours with αCD3/αCD28 (both 0.5 µg/mL) and activated CD4 and CD8 T cells were sorted for analysis by quantitative proteomics. Data was analysed using proteomic ruler method (Wisniewski et al., 2014) to estimate protein copy number per cell. An interactive version of the proteomics expression data is available for exploration on the Immunological Proteome Resource website: immpres.co.uk (G) Total protein content (µg/million cells) (one-way ANOVA), (H) Volcano plots of p-value (-log10) versus fold-change (log2) in protein copy number between Pim dKO and WT. Horizontal dotted line represents multi-test correction cut-off of q=0.05, vertical dotted line shows 1.5-fold change. Phosphoribosyl Pyrophosphate synthase 1 like 1 (Prps1l1), was found to be higher in Pim dKO CD8 T cells, but was a low confidence quantification (based on only 2 unique peptides) with no known function in T cells. Lymph node single cell suspensions from WT and Pim dKO mice were labelled with CellTrace Violet (CTV) and (I) cells were cultured in IL-7 (5ng/mL) +/- rapamycin (20 nM) and CD8 T cell numbers measured over time or (J) cells were activated with αCD3/αCD28 (both 0.5 µg/mL) +/- rapamycin (20 nM) and CD8 T cell mean division number was calculated over time (two-way ANOVA).

Symbols in bar charts represent biological replicates, symbols in (I) represent the mean. Error bars show mean ± S.D. Flow cytometry dot plots and histograms are representative of (D-E) n = 3, except for IFNγ staining which is n=2 or show pooled data from (I) n=3-4 and (F, J) n=5 biological replicates, with data collected over at least 2 independent experiments. Quantitative proteomics was performed on biological triplicates.

PIM1 and PIM2 have similar substrate selectivity and hence functional redundancy (Mikkers et al., 2004; Bullock et al., 2005; Fox et al., 2005; Peng et al., 2007; Warfel and Kraft, 2015). Accordingly, to explore the role of the PIM kinases in T cells, mice with a whole body deletion of Pim1 and Pim2, Pim1-/- (Pim1 KO) (Laird et al., 1993) and Pim2-/- or Pim2-/Y (Pim2 KO) (Mikkers et al., 2004), were backcrossed for >10 generations onto a C57BL/6 background and inter-crossed to generate Pim1/Pim2 double knockout (Pim dKO) mice.

There were normal proportions of peripheral T cells in spleens of Pim dKO mice (Fig S1A) similar to what has been reported previously in Pim dKO mice on an FVB/N genetic background (Mikkers et al., 2004), though the total number of T cells and splenocytes was lower than in age/sex matched wild-type (WT) mouse spleens (Fig S1B-C). This was not attributable to any one cell type (Fig S1A)(James et al., 2021) but was instead likely the result of these mice being smaller in size, a phenotype that has previously been reported in Pim1/2/3 triple KO mice (Mikkers et al., 2004).

We tested how Pim dKO T cells respond to immune activation and observed a normal increase in cell size in response to in vitro immune activation with αCD3/αCD28 agonist antibodies (Fig 1D). 24-hour αCD3/αCD28 activated Pim dKO CD4 and CD8 T cells also normally upregulated expression of established markers of T cell activation including CD25, CD44 and the transferrin receptor CD71 (Fig 1E). 24-hour TCR-activated Pim dKO CD8 T cells also produced normal levels of the key effector cytokine IFNγ (Fig 1E) and there was no discernible difference in TCR-driven proliferation between WT and Pim dKO T cells over 3 days in culture (Fig 1F, J). To systematically explore in an unbiased way if Pim1/Pim2- deficiency impacted TCR-driven activation phenotypes, we performed high-resolution mass spectrometry to measure the proteomes of 24 hour αCD3/αCD28 activated WT and Pim dKO CD4 and CD8 T cells. Proteomics analysis confirmed that no catalytically active PIM1 and PIM2 protein were made in Pim dKO mice (Fig S2). These experiments quantified expression of >7000 proteins but found no substantial quantitative or qualitative differences in protein content or proteome composition in activated WT versus Pim dKO CD4 and CD8 T cells (Fig 1G-H) (Table S1). Collectively these results indicate that PIM kinases do not play an important unique role in the signalling pathways used by the TCR and CD28 to control T cell activation.

One prominent study has reported that Pim1/Pim2 deficiency sensitises T cells to the mTORc1 inhibitor rapamycin (Fox et al., 2005), which suggests mTORc1 can substitute for the PIM kinases to control T cell activation. This previous study showed Pim dKO, but not WT, naïve T cells cultured in IL-7 died in response to rapamycin treatment. It was also shown that rapamycin synergised with Pim1/Pim2-deficiency to substantially reduce TCR-driven activation and proliferation (Fox et al., 2005). However, we found no impact of Pim1/Pim2- deficiency on the ability of naïve CD8 T cells cultured with IL-7 to survive in response to rapamycin treatment (Fig 1I). Furthermore, we found no difference in the effect of rapamycin on mean division number in αCD3/αCD28 activated Pim dKO versus WT CD8 T cells (Fig 1J). Collectively these data show that PIM 1 and PIM2 are rapidly and strongly expressed in response to immune activation, but they are dispensable for the antigen receptor and CD28-driven proteome remodeling that initiates T cell clonal expansion and differentiation. We also found no evidence that mTORc1 was substituting for Pim1 and Pim2 loss.

PIM kinases do not control IL-15-driven CD8 T cell memory differentiation

The proliferative expansion and differentiation programs of antigen activated CD8 T cells are shaped by cytokines. In this context, PIM1 and PIM2 kinase expression is regulated by the cytokines IL-2 and IL-15 which are key to shaping CD8 T cell effector and memory cell differentiation respectively (Weninger et al., 2001; Cornish et al., 2006; Buck et al., 2016).

Both IL-2 and IL-15 upregulate PIM1 and PIM2 expression in antigen primed CD8 T cells (Fig 2A) and are required to sustain PIM1 and PIM2 expression (Rollings et al., 2018).

Pim1/Pim2 deficiency reduces IL-15-driven CD8 T cell proliferation but does not prevent memory differentiation.

(A) OT1 lymph node cell suspensions were SIINFEKL peptide activated for 36 hours, washed then cultured with no cytokine, IL-15 (20 ng/mL) or IL-2 (20 ng/mL) for 4 or 24 hours. Western blots of PIM1 (two isoforms of 44 and 34 kDa, non-specific band indicated by *), PIM2 (three isoforms of 40, 37 and 34 kDa) or pSTAT5 Y694 expression. (B) Schematic of cytokine driven memory and effector CD8 T cell expansion and differentiation cultures. Lymph node or spleen cell suspensions were activated for 2 days with TCR stimulus + cytokine, washed, then split daily into fresh media + cytokine. (C) WT (Ly5.1) and Pim dKO LN suspensions were mixed at a 50:50 ratio for T cells and cultured as outlined in (B) with TCR stimulus αCD3/αCD28 (both 0.5 µg/mL) + cytokine IL-15 (20 ng/mL), and CD8 T cell number was measured daily. (D) WT and Pim dKO T cells were expanded with IL-15 in separate cultures as per (B-C) and % live cells (PI-ve) were assessed on days 4 and 6 (two-way ANOVA). (E-J) WT and Pim dKO CD8 T cells were activated with TCR stimulus αCD3/αCD28 (both 0.5 µg/mL) + cytokine IL-15 (20 ng/mL), expanded with IL-15 as per (B), with an additional CD4 T cell magnetic depletion step on day 3 of culture. CD8 T cells were harvested on day 6 for parallel RNAseq and proteomic analysis. An interactive version of the proteomics expression data is available for exploration on the Immunological Proteome Resource website: immpres.co.uk (E) Fold- change in mRNA expression between Pim dKO and WT versus average mRNA expression (TPM). mRNA expression (Transcripts per million, TPM) of (F) secondary lymphoid homing receptors Sell, Ccr7, S1pr1 and (G) key transcription factors involved in CD8 T cell memory differentiation and maintenance Tcf7, Klf2, Foxo1, Foxo3, Id3. (H) WT vs Pim dKO protein copy numbers, differentially expression proteins (FC>1.5, q<0.05) are highlighted in red (I) Protein copy numbers per cell for key mitochondrial proteins DRP1, OPA1 and CPT1A. (J) WT vs Pim dKO protein copy numbers, mitochondrial proteins (as defined in MitoCarta 3.0) are highlight in pink. Symbols in bar charts represent biological replicates, symbols in (C,E,H,J) represent the mean. Error bars show mean ± S.D. Data are representative of (A) n=3 or show pooled data from (C) n = 4, and (D) n=5 biological replicates with data collected over at least 2 independent experiments. Quantitative proteomics and RNAseq was performed on biological triplicates. ** q≤0.01, fold-change (FC) shown on bar graphs when q<0.05.

Accordingly, we assessed the contribution of PIM1 and PIM2 kinases to cytokine-driven expansion and differentiation of antigen activated CD8 T cells. In these experiments, antigen-primed CD8 T cells are expanded in IL-15 to generate memory phenotype CD8 T cells or expanded in IL-2 to generate effector cytotoxic T lymphocytes (CTL) (Weninger et al., 2001; Cornish et al., 2006; Buck et al., 2016) (Fig 2B).

Previous work has shown that PIM1 and PIM2 mediate IL-15 induced proliferation of gut intra-epithelial lymphocytes (IEL) (James et al., 2021). However, their role in IL-15 responses in conventional antigen receptor activated CD8 T cells is not known. We therefore first examined the impact of Pim1/Pim2-deficiency on IL-15-driven conventional CD8 T cell expansion. Antigen experienced WT CD8 T cells show a robust proliferative response to IL- 15 such that there is an ∼90-fold expansion of the CD8 T cells between days 2 and 6 of culture (Fig 2C). Immune activated Pim dKO CD8 T cells also proliferated strongly in response to IL-15, but showed a moderate proliferative disadvantage resulting in a ∼4-fold difference in total cell number by day 6 of the culture (Fig 2C). There was no impact of Pim1/Pim2-deficiency on the ability of IL-15 to sustain CD8 T cell viability (Fig 2D). Antigen- primed CD8 T cells expand in IL-15 to generate memory phenotype CD8 T cells (Weninger et al., 2001). We therefore examined whether Pim1/Pim2-deficiency had any impact on the ability of IL-15 to induce memory differentiation. Accordingly, we performed parallel bulk RNAseq and high-resolution mass spectrometry to analyse the transcriptome and proteome of IL-15 expanded WT and Pim dKO CD8 T cells. The RNAseq analysis quantified ∼14,000 unique polyA+ mRNA and using a cut off of >1.5 fold-change and q-value <0.05 we saw that the abundance of 381 polyA+ RNA was modified by Pim1/Pim2-deficiency (Fig 2E) (Table S2A). Of the 381 RNA that were detected as differentially expressed (FC>1.5, q<0.05) only ∼30% were detected with an average expression over 10 transcripts per million (TPM), of these gene ∼36% were non-coding RNA, uncharacterised genes or pseudogenes, leaving only 72 substantially expressed protein coding mRNA that were differentially expressed in Pim dKO versus WT IL-15 expanded CD8 T cells (Table S2B). These data indicate that Pim1/Pim2-deficiency has very little effect on IL-15 induced protein-coding mRNA expression in differentiated T cells. We then specifically examined if Pim dKO T cells had any differences in mRNA that are known to be critical for memory cell differentiation. We found that Pim1/Pim2-deficiency did not decrease expression of mRNA levels for critical memory T cell genes such as the adhesion molecules and chemokine receptors, Sell (CD62L), Ccr7 and S1pr1 that direct the trafficking of CD8 T cells to secondary lymphoid tissue. (Fig 2F).

Pim1/Pim2-deficiency also did not reduce expression of mRNA for the key memory cell transcription factors Tcf7, Foxo1, Foxo3 Klf2 and Id3. (Fig 2G). Indeed, if anything, there were small increases in expression of the Klf2, Id3 and Ccr7 mRNA in Pim dKO versus WT IL- 15 expanded CD8 T cells (Fig 2F-G).

Critically, when the proteomes of IL-15 expanded WT and Pim dKO CD8 T cells were analysed and expression of ∼6900 total proteins quantified, there were no major changes in the proteomes between WT and Pim dKO IL-15 expanded CD8 T cells (Fig 2H, Table S3). Of note, the mitochondrial proteome composition of IL-15 expanded WT and Pim dKO CD8 T cells were indistinguishable (Fig 2I-J). This is pertinent because one key feature of how IL-15 controls memory T cells is via mitochondrial remodelling to support energy production via oxidative phosphorylation and fatty acid oxidation (Buck et al., 2016) and it has been described that PIM kinases can control mitochondrial phenotypes in cancer cell lines and cardiomyocytes (Din et al., 2013; Chauhan et al., 2020). These data show that PIM1 and PIM2 do not drive IL-15-mediated metabolic or differentiation programs in antigen-primed CD8 T cells.

PIM kinases selectively modulate IL-2 controlled effector CD8 T cell differentiation

IL-2 promotes the expression of PIM1 and PIM2 kinases (Fig 2A) and in this context, we noted that the copy numbers per cell of PIM1 protein is considerably higher in actively expanding antigen-activated CD8 cells maintained with IL-2 compared to IL-15 (Fig 3A). We therefore examined the role of PIM1 and PIM2 in the IL-2 signalling pathways that control the differentiation of effector CTL. The data show that Pim dKO antigen activated CD8 T cells responded normally to IL-2 to induce STAT5 Y694 phosphorylation (Fig 3B). They also express normal levels of the IL-2-receptor alpha chain, CD25, (Fig 3C), which is sustained by STAT5. Antigen activated WT CD8 T cells have a strong survival and proliferative response to IL-2 producing an ∼270-fold expansion between days 2 and 6 of culture in saturating IL-2 concentrations (Fig 3D). There was no impact of Pim1/Pim2-deficiency on the ability of IL-2 to maintain CD8 T cell viability (Fig 3E) and Pim dKO cells CD8 T cells cultured in IL-2 proliferated robustly. They did however exhibit a very mild proliferative disadvantage resulting in a ∼2-fold difference in total cell number by day 6 of the culture (Fig 3D). We then examined whether Pim1/Pim2-deficiency had any impact on the differentiation of Pim dKO CD8 T cells into effector CTL. Expression of activation markers CD44 and CD71 in IL-2- expanded CTL were similar between WT and Pim dKO CD8 T cells on days 3 and 6 of co- cultures (Fig 3F). However, one consistent difference between IL-2 maintained WT and Pim dKO CD8 T cells is that WT CTL downregulated expression of CD62L (L-selectin, Sell), whereas the Pim dKO CTL sustained higher CD62L expression (Fig 3G). We also noted that IL-2 maintained Pim dKO CTL had reduced forward and side scatter properties i.e were smaller and less granular than corresponding WT CTL (Fig 3F).

Pim dKO IL-2 differentiated effector T cells have reduced cell size and sustained expression of CD62L

(A) Estimated copies per cell of PIM1 and PIM2 protein from published quantitative proteomics analysis (Howden et al., 2019; Brenes et al., 2023) of CD8 T cells expanded in IL-2 or IL-15 as outlined in Fig 2B. (B-D, F-G) WT (Ly5.1) and Pim dKO lymph node or spleen single cell suspensions were mixed at a 50:50 ratio of T cells, activated for 2 days with αCD3/αCD28 (both 0.5 µg/mL) and IL-2 (20 ng/mL), washed then split into fresh medium containing IL-2 (20 ng/mL) daily (as per Fig 2B). Some of the mixed cell suspensions were also cultured in IL-7 (5 ng/mL) to sustain a naïve T cell reference. (B) WT and Pim dKO CTL were treated 1 hour +/- Jak1/3 inhibitor Tofacitinib (100 nM) (negative control) before pSTAT5 Y694 expression was measured on day 3 and 6 of culture, (C) surface CD25 expression was measured on days 3 and 6 or culture, (D) CD8 T cell number vs time was calculated, (F) CD8 T cell FSC-A, SSC-A and surface activation markers (CD44, CD71) were measured on days 3 and 6 of culture (G) expression of adhesion molecule CD62L was measured daily. (E) WT and Pim dKO T cells were activated and expanded with IL-2 as per (Fig 3B-D, F-G) except in separate cultures and % live cells (PI-ve) was assessed on days 4 and 6 (two-way ANOVA). Symbols in bar charts represent biological replicates, symbols in (D) represent the mean. Error bars show mean ± S.D. Data are representative of (B, G) n=4, (C,F) n=6 or show pooled data from (D) n=4, (E) n=6 biological replicates with data collected over at least 2 independent experiments

To explore the apparent size change between Pim dKO and WT IL-2 expanded CTL, we used high resolution mass spectrometry to compare the proteomes of day 6 IL-2 cultured WT and Pim dKO CTL. These analyses quantified expression of ∼6,900 proteins and showed a small decrease in the total protein mass of IL-2 maintained Pim dKO CD8 T cells compared to WT cells (Fig 4A, Table S4). The data showed down-regulation of the expression of 513 proteins (>1.5 fold-change, q<0.05) in Pim dKO CTL along with increased copy numbers of 33 proteins (Fig 4B, Table S4). The transcription factor profile of Pim dKO T cells was consistent with that of an effector CTL with high levels of TBX21 and low levels of TCF1 (Tcf7) (Fig 4C). Pim dKO T cells also exhibited an effector-like mitochondrial proteome profile (Fig S3). However, there were salient differences in proteins important for effector functionality between WT and Pim dKO CD8 T cells. These included lower levels of the glucose transporters, SLC2A1 and SLC2A3 (GLUT1 and GLUT3) (Fig 4D) which are the major glucose transporters in CD8 T cells (Hukelmann et al., 2016) and are key for enabling T cells to fuel high rates of glycolysis and protein O-GlcNAcylation required to support effector functions (Chang et al., 2015; Swamy et al., 2016). Pim dKO CTL also had decreased expression of a number of proteins involved in fatty acid and cholesterol biosynthesis including SCD2, HMGCS1, MVK, IDI1, FDPS, FDFT1 and SQLE (Fig 4E). Another striking observation was that the Pim dKO CD8 cells had reduced levels of a number of key effector proteins including multiple granzymes, perforin, IFNγ and TNFα (Fig 4F-H). Orthogonal flow cytometric analysis of Granzyme B and IFNγ levels confirmed reduced levels of these effector molecule in IL-2 expanded Pim dKO CTL, though IFNγ protein is only lowly produced in IL-2 maintained CTL (Fig 4I). In parallel experiments we also examined the impact of an acute loss of PIM kinase activity on the proteomes of IL-2 maintained CTL to see if this could recapitulate key observations from Pim dKO T cells. When IL-2 maintained WT CTL were treated for 24 hours with the pan-PIM kinases inhibitors PIM447 or AZD1208 we saw reductions in cell proliferation, cell size, cell granularity, fatty acid biosynthesis proteins SCD1-3, glucose transporters SLC2A1, SLC2A3 and a reduction in Granzyme B expression (Fig S4A-E, Table S5). These experiments corroborated some key observations from the experiments with Pim dKO CTL, where Pim1 and Pim2 had been deleted over the lifespan of the T cell.

Major glucose transporters and effector proteins are reduced in Pim dKO IL-2 expanded CTL

WT and Pim dKO CD8 T cells were activated for 2 days with αCD3/αCD28 (both 0.5 µg/mL) and IL-2 (20 ng/mL), washed then split into fresh medium containing IL-2 (20 ng/mL) daily (as per Fig 2B), with an additional CD4 T cell magnetic depletion step on day 3 of culture. CD8 T cells were harvested on day 6 of culture for high resolution mass spectrometry. An interactive version of the proteomics expression data is available for exploration on the Immunological Proteome Resource website: immpres.co.uk (A) Estimated total protein content per cell (student t-test). (B) Volcano plots Pim dKO vs WT protein copy numbers, differentially expressed proteins (FC>1.5, q<0.05) are highlighted in red. Estimated protein copy number per cell of (C) transcription factor TBX21 and TCF1 (D) glucose transporters SLC2A1 and SLC2A3. (E) Volcano plots Pim dKO vs WT protein copy numbers. Proteins with KEGG term = “terpenoid backbone biosynthesis”, “biosynthesis of unsaturated fatty acids” or “steroid biosynthesis” are highlighted with proteins with FC>1.5, q<0.05 shown in red and proteins with FC<1.5 and/or q>0.05 shown in pink. (F) Heatmap of protein copy numbers for granzymes, perforin and effector cytokines. Estimated protein copies for (G) major cytolytic Granzymes A and B and (H) IFNγ. (I) Granzyme B and IFNγ expression was measured by flow cytometry in day 6 IL-2 expanded WT and Pim dKO CTL.

Symbols in bar charts show biological replicates. Error bars show mean ± S.D. Data are representative of (G) n = 3-4, with data collected over at least 2 independent experiment Quantitative proteomics was performed on biological triplicates. * indicates q<0.05, fold-change (FC) shown on graph when q<0.05

To examine the molecular regulation underpinning PIM1 and PIM2 kinase control of key metabolic and effector proteins we also performed RNAseq on IL-2 differentiated WT and Pim dKO CTL samples collected in parallel with the proteomics analysis described in Fig 4. In these experiments ∼14,000 polyA+ RNA were quantified and using a cut off >1.5 fold- change, q<0.05 we saw that the abundance of 223 mRNA were decreased and 155 increased by Pim1/Pim2 deficiency (Fig 5A, Table S6). These data revealed that the reduced expression of perforin and Granzymes C-G and K protein corresponded to decreases in their mRNA (Fig 5B-C). However, there were striking examples where a clear decrease in protein levels did not correspond with an appreciable decrease in mRNA expression. These included the predominant cytolytic Granzymes B and A (Fig 5C-D), as well as both glucose transporters SLC2A1 and SLC2A3 (Fig 5E). Furthermore, contrasting differential expression of proteins and mRNA data in Pim dKO vs WT CTL revealed that ∼75 % of the proteins whose expression was changed in Pim dKO T cells (FC>1.5, q<0.05), exhibited no strong corresponding changes in their mRNA (FC<1.2) (Fig 5F, Table S7). Indeed, of the ∼500 proteins whose expression decreased at the protein level in Pim dKO CTL only 17 showed a strong decrease in magnitude at the mRNA level as well (Fig 5G, Table S7).

Disconnect between protein and mRNA expression in Pim1/Pim2-deficient effector CTL corresponds with a reduction in protein synthesis

RNAseq analysis was performed in day 6 IL-2 expanded WT and Pim dKO CD8 T cells which were collected in parallel with proteomics analysis described in Fig 4A. (A) Volcano plot of RNAseq data, differentially expressed mRNA (FC>1.5, q<0.05) are highlighted in red. (B) Volcano plot of RNAseq data, Granzymes C-K, perforin, Pdcd4 and Sell are highlighted in red. (C) Heatmap of mRNA expression (TPM) for granzymes, perforin and effector cytokines. Bar chart of mRNA expression (TPM) of (D) Granzymes A and B (E) Glucose transporters Slc2a1 and Slc2a3. (F-G) Fold change of PimdKO/WT protein (copies) from proteomics analysis described in Fig 4 vs mRNA (TPM) (F) highlighting in red proteins that are differentially expressed (FC>1.5, q<0.05) where mRNA is not substantially different (FC<1.2) and (G) highlighting in red protein and mRNA that are both differentially expressed (FC>1.5, q<0.05). (H) Estimated cytosolic ribosome content per cell (left), % ribosome of total cellular protein content (right). (I) Estimated protein copy number per cell of translation repressor PDCD4. (J) PDCD4 expression measured by flow cytometry on day 3 and 6 in IL-2 expanded WT vs Pim dKO CD8 T cells. (K) Estimated protein copy number per cell of EIF4A1. (L) adjusted ratio of PDCD4 : EIF4A1 (assuming 1 PDCD4 binds 2x EIF4A1) in WT and Pim dKO proteomes. (M) Protein synthesis measured by OPP incorporation in day 6 IL-2-expanded WT CTL treated for 24 hours with pan PIM kinase inhibitors PIM447 (5 µM) or AZD1208 (10 µM). 30 min cycloheximide (100 µg/mL) treatment gives no protein synthesis background control.

Symbols in bar charts show biological replicates: error bars show mean ± S.D. Data are representative of (M) n=2 biological replicates collected over two independent experiments, (J) n = 2 biological replicates. Quantitative proteomics and RNAseq were performed on biological triplicates. * indicates q<0.05, fold-change (FC) shown on graph when q<0.05.

These observations could reflect that the PIM kinases are required for IL-2 to maximally control protein synthesis in CD8 T cells. To examine this hypothesis we first interrogated our proteomics and transcriptomics data. Consistent with this hypothesis we observed a small reduction in ribosome protein content in Pim dKO CTL, which scaled with the reduction in total protein content (Fig 5H). Moreover, the translational repressor PDCD4 was increased in Pim dKO CTL (Fig 5I). This could be confirmed by flow cytometry (Fig 5J) and was underpinned at the mRNA level (Fig 5B). Similarly, 24 hour inhibition with pan PIM kinases inhibitors was sufficient to substantially increase PDCD4 protein expression (Fig S4F). We also noted that there was a small decrease in the expression of EIF4A1, a key component of the eIF4F translation initiation complex, in Pim dKO IL-2 maintained CTL (Fig 5K). PDCD4 inhibits translation by binding EIF4A1 in a 2:1 ratio, preventing its interaction with the EIF4F complex (Suzuki et al., 2008). Together the changes in both proteins equated to an approximate halving of the ratio between EIF4A1 and PDCD4 (Fig 5L). These data all support that loss of PIM kinases could reduce protein synthesis in IL-2 maintained CTL. To test this hypothesis we used a single cell assay that quantifies the incorporation of an analogue of puromycin (OPP) into newly synthesized protein chains in the ribosome to assess cellular rates of protein synthesis. We found that 24 hours of treatment with pan-PIM kinase inhibitors PIM447 or AZD1208 caused a quantitative reduction in protein synthesis rates of IL-2 expanded CTL (Fig 5M). This data is consistent with PIM kinase effects on protein expression being via control of protein translation.

PIM1 and PIM2 regulate mTORc1 and sustain lymphoid homing in Pim1/Pim2-deficient effector CTL

The pattern of proteome remodelling in Pim dKO CTL had some phenotypic features in common with the previously reported changes that occur following inhibition of the Serine/Threonine kinase mTORc1 in CTL (Hukelmann et al., 2016; Howden et al., 2019). These included: increased expression of PDCD4 and decreased expression of cholesterol biosynthesis enzymes and effector molecules including granzymes, perforin, IFNγ and TNFα (Hukelmann et al., 2016; Howden et al., 2019). In this respect, mTORc1 activity in CTL is acutely sensitive to reduced availability of glucose (Rolf et al., 2013) and the reduced expression of glucose transporters in Pim dKO T cells might predict that these cells would have lower activity of mTORc1. We therefore assessed if PIM kinases were required for IL-2 activation of mTORc1. In these experiments we compared the phosphorylation of S6K1 (p70 S6-kinase 1) on its mTORC1 substrate sequence T389 in WT and Pim dKO CD8 T cells by western blot. These data show a modest reduction of mTORc1 activity in the Pim dKO T cells (Fig 6A). We also assessed the activity of S6K1 at a single cell level by quantifying the phosphorylation of its downstream target protein S6 ribosomal protein on S235/S236 by flow cytometry over the course of IL-2 driven CTL expansion. These data also show that Pim1/Pim2-deficiency reduced IL-2 induced mTORc1 dependent phosphorylation of S6 but did not compromise the ability of the Pim dKO cells to phosphorylate S6 in response to phorbol esters and calcium ionophores; pharmacological activators that induce RSK mediated S6 phosphorylation (Fig 6B).

PIM kinases regulate mTORc1 activity and lymphoid homing in effector CTL

(A) Western blot of pS6K T389 and pan S6K from day 6 WT and PimdKO IL-2 CTL (paired student t-test, *= p<0.05). (B) WT and Pim dKO T cells were mixed in a 50:50 ratio, activated with αCD3/αCD28 (both 0.5 µg/mL) and IL-2 (20 ng/mL) and expanded in IL-2 as per Fig 2B and pS6 (Ser235/236) measured after 1 hr +/- PdBU and ionomycin (positive control), +/- rapamycin (mTORc1 inhibitor, negative control) or no additional treatment. mRNA expression (TPM) from RNAseq analysis of IL-2 expanded WT and Pim dKO CTL described in Fig 5A for (C) cell homing receptors S1pr1, Ccr7 and Sell and (D) transcription factor Klf2. (E) WT and Pim dKO T cells were activated and expanded with IL-2 in separate cultures as per Fig 2B. On day 6 of culture WT and Pim dKO CTL were labelled with CFSE or CTV, mixed at a 50:50 ratio and transferred into C57BL/6 recipient mice. Values indicate percentage of transferred cells detected in blood, lymph node or spleen 4 or 23 hours post-transfer that were WT or Pim dKO (one-way ANOVA).

Symbols show biological replicates. Error bars show mean ± S.D. Data are representative of (A) n= 6, (B) n=2-4 collected across at least 2 independent experiments and (E) n= 6 recipient mice, from n=2 biological donor replicates. RNAseq was performed on biological triplicates. * q<0.05, ** q<0.01, *** q<0.001, fold-change (FC) shown on graph when q<0.05 but FC<1.5.

Previously, it has been shown that decreased glucose availability or inhibition of mTORc1 prevents the downregulation of CD62L that normally accompanies IL-2 induced CTL differentiation (Sinclair et al., 2008; Finlay et al., 2012). Thus, the reduced expression of glucose transporters and reduced activity of mTORc1 in Pim dKO IL-2 expanded CTL provides a potential explanation for the striking phenotype shown in Fig 3G, that IL-2 maintained Pim dKO CTL retain high levels of CD62L/L-selectin. The IL-2 CTL transcriptomics data showed there is increased expression of CD62L at the mRNA level in IL-2 maintained PimdKO CTL (Fig 5B, 6C). The expression of Sell (CD62L) mRNA is controlled by the transcription factor KLF2 and the expression of Klf2 mRNA is also increased in IL-2 maintained Pim dKO CTL (Fig 6D) as are mRNA levels for other KLF2 targets: the chemokine receptors Ccr7 and S1pr1 (Fig 6C).

CD62L/L-selectin controls T cell adhesion to the endothelium of high endothelial venules and is essential for lymphocyte transit from blood into secondary lymphoid tissue such as lymph nodes. CCR7, and the S1P1 receptor also direct migration by controlling T cell entry and egress respectively into secondary lymphoid tissue. In this context, the loss of CD62L, CCR7 and S1PR1 expression following immune activation reprograms the trafficking of effector T cells away from lymphoid tissue toward sites of inflammation in peripheral tissues (Sinclair et al., 2008; Nolz et al., 2011). Accordingly, the differences between Pim dKO and WT CTL in terms of CD62L expression could indicate that in addition to having reduced effector protein expression, Pim dKO CTL also do not switch their trafficking behaviour to that of a fully differentiated effector CTL and rather retain the capacity to home to lymphoid tissues. To test this possibility in vivo, we performed adoptive transfer experiments comparing the ability of WT effector CTL and Pim dKO CTL to home to secondary lymphoid tissues. In these experiments, αCD3/αCD28 activated WT or Pim dKO T cells were differentiated in IL-2 and cells then were labelled with either CFSE or CTV dyes; mixed at a ratio of 1:1, and adoptively transferred into C57BL/6 hosts. Fig 6E shows that Pim dKO CTL, but not WT CTL, retained substantial capacity to home to secondary lymphoid organs and, hence, accumulated in lymph nodes and spleen: PIM kinases are thus required for the normal reprogramming of CTL trafficking.

Discussion

The objective of the present study was to use quantitative analysis of T cell proteomes and transcriptomes to explore how deletion of the Serine/Threonine kinases PIM1 and PIM2 impacts CD8 T cell activation and differentiation. The expression of PIM1 and PIM2 is induced during immune activation of CD8 T cells by antigen and co-stimulation and then sustained to varying degrees by the cytokines IL-2 or IL-15. The loss of PIM1 and PIM2 was shown to have no effect on the CD8 T cell programs controlled by antigen receptor/co-stimulation and had very little impact on the IL-15-driven transcriptional and proteomic programs in memory differentiated CD8 T cells. However, PIM1 and PIM2 were required for IL-2 programming of effector CD8 T cell differentiation. IL-2 controls essential cytolytic effector and metabolic programs in CTL, is a key cytokine for CD8 T cell anti-viral immunity and is also used to produce cytotoxic T cells for anti-cancer immunotherapy (Kalia and Sarkar, 2018). The understanding that PIM kinases mediate IL-2 control of the expression of glucose transporters, lipid metabolic enzymes and key cytolytic proteins such as granzymes A and B in effector CD8 T cells is thus fundamental information about how this key cytokine controls T cells. It was also important to understand that PIM kinases did not control the transcriptional programs that direct expression of glucose transporters and effector enzymes and the hence the full consequences of PIM kinase deletion could only be assessed by in depth analysis of the proteomes of Pim1/Pim2-deficient CTL and not from mRNA sequencing. The loss of PIM kinase activity reduced protein synthesis rates in CTL which explains the discordance between mRNA and protein reported herein. The reduction in protein synthesis was modest but previous studies have shown that modest reductions in protein synthesis and loss of mTORc1 activity impact CD8 T cell differentiation programs (Cornish et al., 2006; Hukelmann et al., 2016; Howden et al., 2019). PIM kinase regulation of protein translation has been reported previously in mouse B cell lymphoma and mouse embryonic fibroblasts (Schatz et al., 2011) and some of the key examples of proteins we found to be unaltered at the mRNA level but down-regulated at the protein level have prior reports of sensitivity to translational regulation in immune cells (Fehniger et al., 2007; Ricciardi et al., 2018). Reduced protein synthesis capacity in cells lacking PIM kinases could be caused by reduced mTORc1 activity combined with increased expression at the mRNA and protein level of the translational repressor PDCD4. Moreover, protein synthesis is highly energetically demanding and once CTL have reduced glucose transporter expression this would reduce glucose availability to fuel protein production. These data collectively point towards modulation of protein translation as an important target of PIM kinase activity downstream of IL-2 in CD8 T cells.

One other important insight from the current work is that PIM kinases are required for IL-2 to fully repress expression of the adhesion molecule CD62L and the chemokine receptors CCR7 and S1PR1. These molecules play a key role in controlling T cell homing to secondary lymphoid tissue and are expressed at high levels in naïve and memory T cells. IL-2 induces down regulation of CD62L, CCR7 and S1PR1 expression as part of a program that causes effector cells to lose their capacity to home to secondary lymphoid tissues (Sinclair et al., 2008; Nolz et al., 2011). The present data show that PIM kinases can mediate this IL-2 repression of lymph node homing receptors and the ability of IL-2 to switch off the naïve T cell trafficking program. IL-2 maintained Pim dKO CD8 effector T cells thus retain the capacity to home to secondary lymphoid tissues in vivo. This lymph-node homing pattern is also a memory CD8 T cell phenotype but here it is pertinent that the current in-depth analysis of the transcriptomes and proteomes of Pim1/Pim2-deficient IL-2 differentiated CD8 T cells show that the loss of PIM kinases does not cause effector T cells to differentiate instead into memory cells. For example, IL-2 maintained Pim dKO CD8 T cells do not re-express key molecules that maintain the stem-like properties of memory cells such as the transcription factor TCF7 and IL-7 receptor (Table S2, S6). Nor do Pim dKO CTL cultured in IL-2 have the mitochondrial proteome of a memory T cell and PCA plots of IL-15 and IL-2 proteomics and RNAseq data show that Pim dKO IL-2 expanded CTL are still much more similar to IL-2 expanded WT CTL than to IL-15 expanded CTL (Fig S5). Hence Pim1/Pim2-deficiency prevents optimal differentiation of effector CTL but does not cause these cells to fully switch to a memory phenotype. This is worth knowing as pan-PIM kinase inhibitors are in development as anti-cancer drugs and it has been suggested that PIM inhibitors might be used as a strategy to generate more stem-memory-like CD8 T cells in combination with cancer immunotherapy treatments (Chatterjee et al., 2019; Clements and Warfel, 2022). Information about how the loss of PIM kinases will impact CD8 T cells is very relevant to informed usage of such drugs.

Finally, the selectivity of PIM kinases for IL-2 control of conventional CD8 T cell differentiation with no evidence that PIMs were important for IL-15 responses beyond a modest effect on cell proliferation was interesting. This failure to see a role for PIM signalling in conventional peripheral CD8 T cell differentiation is in contrast to observations that PIM1 and PIM2 direct IL-15 control of the metabolic programming of intestinal IEL (James et al., 2021). A notable difference between these scenarios was that in IEL, IL-15 was being used individually to drive immune activation and was sufficient to strongly induce PIM1 and PIM2 expression. In contrast, in our conventional CD8 T cell system we uncovered that IL-15 does not sustain high levels of PIM kinases, which could possibly explain this discrepancy. However, high levels of PIM expression alone are not inevitably predictive of their importance. In this study we observed very high levels of both PIM1 and PIM2 in antigen activated CD8 T cells yet loss of PIM kinases there had no functional impact.

Previously it was suggested that mTORc1 could compensate for PIM kinase deficiency (Fox et al., 2005). However we found no evidence for this, and in fact observed that PIM kinases were required to maximally stimulate mTORc1 activity in IL-2 expanded CTL. Thus, this examination of the role of PIM kinases in multiple signalling situations reveal cell context dependent regulatory roles for PIM kinases rather than hard wired functions that apply to every stimulatory situation in T cells.